---
title: Voronoi Homogeneity Analysis of Categorical Data
author:
  - name: Jan de Leeuw
    orcid: 0000-0003-1420-1797
    email: jan@deleeuwpdx.net
    affiliation:
      - name: University of California Los Angeles
        city: Los Angeles
        state: CA
        url: www.ucla.edu
    license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
  pdf:
    fontsize: 12pt
    include-in-header:
      - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
  html:
    fontsize: 12pt
    include-in-header:
      - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: TBD
---

```{r}
#| label: loadpackages
#| echo: false

suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
suppressPackageStartupMessages(library(MASS, quietly = TRUE))
suppressPackageStartupMessages(library(RSpectra, quietly = TRUE))
suppressPackageStartupMessages(library(deldir, quietly = TRUE))
suppressPackageStartupMessages(library(dismo, quietly = TRUE))
```

```{r}
#| label: sourcecode
#| echo: false

source("data/small.R")
source("data/galo.R")
source("data/cetacea.R")
source("mca.R")
source("mals.R")
source("monotone.R")
source("auxiliaries.R")
source("voronoi.R")
source("plotVHA.R")
```

\sectionbreak 

**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex, html,
pdf, R, and C files are in the public domain. Attribution will be appreciated,
but is not required. The files can be found at
<https://github.com/deleeuw/voronoi>

\sectionbreak

# Introduction

## Homogeneity Analysis

*Homogeneity Analysis (HGA)*, a.k.a. *Multiple Correspondence Analysis (MCA)*,
can be introduced and implemented in many different
ways\footnote{Since PCA and MCA have three-letter
acronyms I added the G of Gifi to the HA of Homogeneity Analysis.}.
We refer to @guttman_41, @tenenhaus_young_85, @bekker_deleeuw_C_88, @gifi_B_90,
@greenacre_blasius_06, @leroux_rouanet_10, @husson_josse_saporta_16, and
@deleeuw_C_23 for a bouquet of discussions and interpretations of the equations
and algorithms defining the HGA technique.

A *variable* is a function that maps on a (finite or infinite) domain of
*objects* into a (finite or infinite) codomain of *values*. In HGA the data are
the values of $m$ variables on the same *carrier*, a finite subset the domain
consisting of $n$ objects. Each variable $j$ has a finite number $k_j$ of
*categories*. The categories constitute the *range* of the variable for the $n$
objects in the carrier. Note that all variables have the same carrier. Each
object maps into a unique category, and each category contains a subset of
objects from the carrier. We also say that a category has a number of objects
that "belong to" the category.

Here is a small example with $n=10$ and $m=3$, taken from @gifi_B_90, page 65.
In the github voronoi repository these data are in data/small.R.

```{r}
#| label: small
#| echo: false

print(small)
```

The key mathematical object in HGA is the *indicator matrix*, a classical way to
code categorical
data\footnote{Since the indicator matrix is the core of the Gifi 
system we avoid the irreverent alternative "dummy coding".}.
For a variable $j$ with $k_j$ categories it is an $n\times k_j$ binary matrix
$G_j$ in which each row $i$ contains a single element equal to one, indicating
into which category the object $i$ maps for variable $j$. The cross product
$G_j'G_j$ is diagonal, with the *marginal frequencies* of the categories on the
diagonal. For two different variables the crossproduct $G_j'G_\ell$ is the
$k_j\times k_\ell$ *cross table* or *contingency table* of the two variables.
Here are the three indicator matrices of our small example, next to each other.

```{r}
#| label: indi
#| echo: false

g1 <- makeIndicator(small[, 1])
g2 <- makeIndicator(small[, 2])
g3 <- makeIndicator(small[, 3])
d1 <- diag(1 / colSums(g1))
d2 <- diag(1 / colSums(g2))
d3 <- diag(1 / colSums(g3))
gg <- noquote(cbind("||", g1, "||", g2, "||", g3, "||"))
colnames(gg) <- c("", "a", "b", "c", "", "p", "q", "r", "", "u", "v", "")
print(gg)
```

We now deviate from the usual Gifi and French approaches
(@husson_josse_saporta_16) and define HGA as a form of non-metric
multidimensional scaling, using the *stress* loss function (@kruskal_64a,
@kruskal_64b). Define\footnote{The symbol $:=$ is used for definitions}
\begin{equation}
\sigma(X,Y_1,\cdots,Y_m,\Delta_1,\cdots,\Delta_m) ):=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}^j(\delta_{i\ell}^j-d(x_i,y_\ell^j))^2,\label{eq-loss1}
\end{equation}
with $w_{i\ell}^j$ given non-negative weights, $d(x_i,y_\ell^j)$ the *Euclidean
distance* between row $i$ of $X$ and row $\ell$ of $Y_j$, and
$\smash{\delta_{i\ell}^j}$ the corresponding *disparity*. 

In classical
HGA there are no weights, i.e. all $w_{i\ell}^j$ in \eqref{eq-loss1}
are equal to one. We have added
weights to the definition of HGA so that, for example, we can easily incorporate
missing data.

Disparities are
required to satisfy the measurement restrictions implied by the data. In HGA
this means that we require that $\smash{\delta_{i\ell}^j=0}$ whenever
$\smash{g_{i\ell}^j=1}$, or, using the elementwise product $\circledast$ of
matries, $\Delta_j\circledast G_j=0$.

HGA minimizes \eqref{eq-loss1} over the $n\times p$ configurations $X$
satisfying $X'X=I$, over the $Y_j$, and over the $\Delta_j$ that satisfy the
measurement restrictions.

Define
\begin{equation}
\sigma_\star(X,Y_1,\cdots,Y_m):=\min_{\Delta_1,\cdots,\Delta_m}\sigma(X,Y_1,\cdots,Y_m,\Delta_1,\cdots,\Delta_m).\label{eq-loss2}
\end{equation}
The minimizing disparity $\delta_{i\ell}^j$ is equal to zero if $g_{i\ell}^j=1$
and equal to $d(x_i,y_\ell^j)$ otherwise. Consequently
\begin{equation}
\sigma_\star(X,Y_1,\cdots,Y_m)=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}^jg_{i\ell}^jd^2(x_i,y_\ell^j).\label{eq-loss3}
\end{equation}
Note that the value of \eqref{eq-loss3} does not depend on the $w_{i\ell}^j$ for
which $g_{i\ell}^j=0$. This implies that we can actually forget about the 
subscript $\ell$ in $w_{i\ell}^j$ and simply use $w_{i}^j$.

We can make a graphical representation of \eqref{eq-loss3}. For variable $j$ we
plot the $n$ points $x_i$ and the $k_j$ points $y_\ell^j$ in a joint plot (or
biplot). We then connect, with straight lines, each category point with the
object points that belong to it. For each category this defines a figure in the
plot which we call a *star*. The *size* of a star is the weighted sum of squares
of the length of its lines. The loss function \eqref{eq-loss3} for variable $j$
is the weighted sum of squares of the length of all $n$ lines in this plot, or
equivalently the sum of the sizes of the $k_j$ stars. The total loss is the sum
of star-sizes over all $m$ plots. Thus HGA aims to place the object-points and
category-points in space in such a way that the stars are as small as possible,
i.e. objects are as close as possible to the categories they belong to.

In matrix notation
\begin{equation}
\sigma_\star(X,Y_1,\cdots,Y_m)=\sum_{j=1}^m \text{tr}\ \{X'W_j X-2\ X'W_jG_jY_j+ Y_j'D_jY_j\}\label{eq-loss4}
\end{equation}
Here $W_j$ is a diagonal matrix with the $w_i^j$ and $D_j$ is the
diagonal matrix of column-sums of $W_jG_j$. Note that in classical HGA, with
$W_j=I$ for all $j$, we have $D_j=G_j'G_j$.

Next define
\begin{equation}
\sigma_{\star\star}(X):=\min_{Y_1,\cdots,Y_m}\sigma_\star(X;Y_1,\cdots,Y_m).\label{eq-loss5}
\end{equation}
The minimum of \eqref{eq-loss4} over the $Y_j$ is attained at
$\hat Y_j:=D_j^{-1}G_j'W_jX$. Category-points are the weighted centroids of the
object-points belonging to the category. This makes stars actually look like
stars. If we substitute these optimal $\smash{\hat Y_j}$ in \eqref{eq-loss3} we
see that $\sigma_{\star\star}(X)$ is indeed equal to the weighted sum of squares of all
lines defining the stars. Minimizing $\sigma_{\star\star}(X)$ over $X'X=I$, and
thus minimizing \eqref{eq-loss1} over $(X, Y_j,\Delta_j)$, defines HGA.


Computing the solution turns out to be a straightforward eigenvalue-eigenvector
problem. The details are in the publications we referenced at the beginning of
this section.

A HGA in two dimensions produces of our small example the following three star
plots. Objects are numbered and in BLUE, categories have bigger labels and are
in RED. Note that there are ten objects, but the sets of objects $(1,4,7)$,
$(8,10)$, and $(6,9)$ have the same *profile* and consequently get mapped into
the same object-point. Thus there are at most six different object-points. The
analysis minimizes the sum of squares of the length of all black lines in the
three plots, under the condition that the matrix of object scores is
column-centered and orthonormal. We can think of the results as a form of *graph
drawing*, where the graph is the union of the $m$ bipartite graphs of all $m$
"belonging to" relations (@deleeuw_michailidis_C_00, @michailidis_deleeuw_A_01).

```{r}
#| label: mca
#| echo: false
#| fig-align: "center"

haux <- mca(list(g1, g2, g3), c(3, 3, 2), 2)
y1 <- haux$ymat[[1]]
y2 <- haux$ymat[[2]]
y3 <- haux$ymat[[3]]
x <- haux$xmat
par(pty = "s")
par(mfrow = c(1, 3))
plot(rbind(x, y1), type = "n", xlab = "", ylab = "")
text(x, as.character(1:10), col = "BLUE", cex = .8)
text(y1, c("a", "b", "c"), col = "RED", cex = 1.5)
for (i in 1:10) {
  if (small[i, 1] == "a") lines(rbind(x[i,], y1[1, ]))
  if (small[i, 1] == "b") lines(rbind(x[i,], y1[2, ]))
  if (small[i, 1] == "c") lines(rbind(x[i,], y1[3, ]))
}
plot(rbind(x, y2), type = "n", xlab = "", ylab = "")
text(x, as.character(1:10), col = "BLUE", cex = .8)
text(y2, c("p", "q", "r"), col = "RED", cex = 1.5)
for (i in 1:10) {
  if (small[i, 2] == "p") lines(rbind(x[i,], y2[1, ]))
  if (small[i, 2] == "q") lines(rbind(x[i,], y2[2, ]))
  if (small[i, 2] == "r") lines(rbind(x[i,], y2[3, ]))
}
plot(rbind(x, y3), type = "n", xlab = "", ylab = "")
text(x, as.character(1:10), col = "BLUE", cex = .8)
text(y3, c("u", "v"), col = "RED", cex = 1.5)
for (i in 1:10) {
  if (small[i, 3] == "u") lines(rbind(x[i,], y3[1, ]))
  if (small[i, 3] == "v") lines(rbind(x[i,], y3[2, ]))
}
```

## Voronoi Homogeneity Analysis

Although using loss function \eqref{eq-loss1} makes HGA a form of non-metric
multidimensional scaling there are some important differences with the more
familiar MDS versions minimizing stress. There is no local minima problem in
HGA, because it turns out to be a symmetric eigenvalue problem. In HGA, unlike
in MDS, we impose the constraint $X'X=I$. Also, and perhaps most importantly, in
HGA we do not really expect the minimum loss function value to be zero, or even
close to zero. The loss function measures deviations from a very unrealistic
"model". Zero loss would imply all object-points coincide with all
category-points they belong to. This can only be realized, except in trivial
cases, if all object and category-points map into a single point, which is
prohibited by the orthonormality constraint on the object scores. Thus we get a
perfectly determinate and stable solution, at low computational cost, but by
imposing strong constraints on the disparities and a rather arbitrary
normalization. We can get around this seemingly unfortunate situation by
admitting we are not fitting or testing a "model", but we are creating "best"
drawings of the bipartite graphs that constitute the data.

The lack of a realistic model makes it interesting to continue use the more
general loss function
\begin{equation}
\sigma(X,Y_1,\cdots,Y_m,\Delta_1,\cdots,\Delta_m) ):=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}^j(\delta_{i\ell}^j-d(x_i,y_\ell^j))^2,\label{eq-loss2}
\end{equation}
with the constraints that if $g_{i\ell}^j=1$ then merely
$\delta_{i\ell}^j\leq\delta_{i\nu}^j$ for all $\nu=1,\cdots,k_j$. This is
significantly weaker than the HGA constraints, which require the $mn$
disparities $\smash{\delta_{i\ell}^j}$ to be zero. If object $i$ belongs to
category $\ell$ of variable $j$ we do no longer require the object-point to
coincide with the category-point, but we merely require that the object-point is
at least as close to category-point $\ell$ as to any of the other $k_j-1$
category-points of that variable.

*Voronoi Homogeneity Analysis* (VHA) of categorical data minimizes loss function
\eqref{eq-loss1} over $X$, over the $Y_j$ and over the $\Delta_j$ satisfying the
relaxed ordinal constraints. It is consequently another form of non-metric
(ordinal) multidimensional scaling. To be more precise, we have $m$ linked
non-metric multidimensional unfolding problems for the binary matrices $G_j$,
which are "linked" because they share the same $X$.

Because of the different ordinal constraints on the disparities, the plots
corresponding with this analysis are different. As in HGA there is a plot for
each variable $j$ with both $n$ object-points and $k_j$ category-points. The
$k_j$ category-points are the *generators* of $k_j$ *Voronoi regions* that
partition the space. Voronoi region $\smash{\mathcal{V}^j_\ell}$ is the region
of space where the points are at least as close to generator $\smash{y_\ell^j}$
as to any of the other generators of variable $j$. And loss \eqref{eq-loss4} is
zero (for a variable) if all object-points are in the "correct" Voronoi region,
i.e. in the Voronoi region of the generator corresponding with the category they
belong to.

One (inefficient, but conceptually easy) way of constructing Voronoi regions in
the plane is to draw the perpendicular bisectors of the (hypothetical) lines
connecting all pairs of category-points of a variable. In higher dimensions
$p>2$ the bisectors are planes of dimension $p-1$. The convex polyhedral regions
thus formed, some bounded and some unbounded, define the Voronoi
regions\footnote{In the plane a Voronoi region is unbounded if and only if its generator is on the boundary of the convex hull of all generators.}.
The literature on properties and computation of Voronoi regions is large, and we
refer to the textbooks of @okabe_boots_sugihara_chiu_00 and
@aurenhammer_klein_lee_13.

Since the category-point itself is obviously in its own Voronoi region and since
Voronoi regions are convex it follows that if we make stars by connecting
category-points with object points as in HGA then we actually require each star
to be in its Voronoi region. It is no longer true that a category-point is the
centroid of points of the objects belonging to the category. The category-point
can even be outside the convex hull of the object-points belonging to the
category.

The constraint $X'X=I$ from HGA is no longer imposed directly, but we somehow
have to deal with the fact that in VHA the minimum of loss function
\eqref{eq-loss1} is always zero. The trivial solution, in which all points and
regions collapse into a single point and all $\Delta_j$ are zero, is always
available and not excluded by the constraints. But, perhaps more urgently, we
have to deal with the types of degenerate solutions familiar from
multidimensional unfolding (@heiser_81, @deleeuw_R_83, @busing_10). For example,
we can collapse all category-points of a variable into a single point and put
all object-points on a sphere with the category-point at its center. Then all
distances between category and object-points are the same and we can choose all
disparities equal to that distance value and obtain loss zero.

We will use some constraints that tend to counteract degeneracy. Also, we always
start our iterations with the HGA analysis, which in most cases already has
small loss. Then we perform iterations until either the distances do not change
from one iteration to the next (up to some epsilon) or the loss becomes zero (up
to some epsilon) or we have reached the maximum number of iterations allowed.

## Spherical Homogeneity Analysis

If $g_{i\ell}^j=1$ and $g_{k\ell}^j=0$ then $d_{i\ell}^j(X,Y)\leq d_{k\ell}^j(X,Y)$.

Thus object-points belonging to a category are in a sphere centered at the
category point, and objects not belonging to that category are outside
the sphere. Thus spheres corresponding with different categories of the
same variable are disjoint. This implies convex hulls are disjoint and
object poinst are in the correct Voronoi region.




\sectionbreak

# Algorithm

We use an *Alternating Least Squares* or *ALS* algorithm, similar to the
approach used in many other non-metric scaling methods, to minimize loss
\eqref{eq-loss5}. For context, see @borg_groenen_05. We alternate minimization
over $\Delta_1,\cdots,\Delta_m$ for given $X$ and $Y_1\cdots Y_m$ with
minimization over $X$ and $Y_1\cdots Y_m$ for given $\Delta_1,\cdots,\Delta_m$.
Using superscript $(\mu)$ for the iteration index we can write for an ALS
iteration
\begin{subequations}
\begin{align}
(\Delta_1^{(\mu)},\cdots,\Delta_m^{(\mu)})&=\mathop{\text{argmin}}_{\Delta_1,\cdots,\Delta_m}\sigma(X^{(\mu)},Y_1^{(\mu)},\cdots,Y_m^{(\mu)},\Delta_1,\cdots,\Delta_m),\label{eq-als1}\\
(X^{(\mu+1)},Y_1^{(\mu+1)},\cdots,Y_m^{(\mu+1)})&=\mathop{\text{argmin}}_{X,Y_1,\cdots,Y_m}\sigma(X,Y_1,\cdots,Y_m,\Delta_1^{(\mu)},\cdots,\Delta_m^{(\mu)}).\label{eq-als2}
\end{align}
\end{subequations}
The first substep \eqref{eq-als1} of iteration $(\mu)$ solves a monotone
regression problem for a simple partial order, in which one of $k_j$ elements
must be smaller than the other $k_j-1$ elements. The second substep
\eqref{eq-als2} solves a metric multidimensional scaling problem. This second
subproblem cannot be solved completely, because it would need an infinite
iterative process by itself. Thus the second substep \eqref{eq-als2} consists of
a finite number of *inner iterations* within the second substep of a single ALS
*outer iteration*.

## Initial Configuration

In all our analysis we use HGA as the initial configuration for both
object-scores and category-scores. The algorithm for HGA is the ALS method used
in @gifi_B_90 to compute the truncated SVD. The repository includes mca.R

## Monotone Regression

The $nm$ monotone regressions are applied for each object $i$ and each variable
$j$. Forgetting about all these indices for the moment we need to minimize the
sum of squares $\|\delta-d\|^2$ over $\delta$, where one of the elements, say
$\delta_\nu$, must be smaller than or equal to all others.

This can be solved by using monotone regression with the primary approach to
ties (@kruskal_64b, @deleeuw_A_77), treating all $\delta_\ell$ with
$\ell\neq\nu$ as ties.

## Majorization

We study the off-diagonal smacof algorithm for MDS, without taking into account
that the loss function \eqref{eq-loss1} comes from $n$ observations on $m$
categorical variables. This improves and generalizes the smacof treatment of the
metric unfolding problem in @heiser_deleeuw_A_79.

We rewrite stress as
\begin{equation}
\sigma(X,Y):=\sum_{i=1}^n\sum_{k=1}^Kw_{ik}(\delta_{ik}-d(x_i,y_k))^2.\label{eq-ploss}
\end{equation}
The distance is
\begin{equation}
d(x_i,y_k)=\sqrt{(X'e_i-Y'e_k)'(X'e_i-Y'e_k)}.
\end{equation}
We will construct a majorization of \eqref{eq-ploss} at $\tilde X$ and
$\tilde Y$. Think of $\tilde X$ and $\tilde Y$ as the current best estimates in
a sequence of iterations. By Cauchy-Schwartz,
\begin{equation}
d(x_i,y_k)\geq\frac{1}{d(\tilde x_i,\tilde y_k)}(X'e_i-Y'e_k)'(\tilde X'e_i-\tilde Y'e_k).
\end{equation}
Define the $n\times K$ matrix-valued function $B(X,Y)$ by
\begin{equation}
b_{ik}(X,Y):=\begin{cases}w_{ik}\frac{\delta_{ik}}{d(x_i,y_k)}&\text{ if }d(x_i,y_k)>0,\\
0&\text{ if }d(x_i,y_k)=0.\end{cases}
\end{equation}
Also define the diagonal matrices $B_{\star}(X,Y)$ with the row-sums of $B(X,Y)$
and $B^{\star}(X,Y)$ with the column-sums.

If we define, as we usually do in smacof,
\begin{subequations}
\begin{equation}
\rho(X,Y):=\sum_{i=1}^n\sum_{k=1}^Kw_{ij}\delta_{ij}d_{ij}(x_i,y_k)
\end{equation}
hen $\rho(X,Y)\geq\rho(X,Y;\tilde X,\tilde Y)$, where
\begin{equation}
\rho(X,Y;\tilde X,\tilde Y):=\text{tr}\ X'B_\star(\tilde X,\tilde Y)\tilde X-\text{tr}\ X'B(\tilde X,\tilde Y)\tilde Y-\text{tr}\ Y'B'(\tilde X, \tilde Y)\tilde X+\text{tr}\ Y'B^\star(\tilde X,\tilde Y)\tilde Y.
\end{equation}
Also define
\begin{equation}
\eta^2(X,Y):=\sum_{i=1}^n\sum_{k=1}^Kw_{ik}\delta_{ij}d_{ij}^2(x_i,y_k)=\text{tr}\ X'W_\star X-2\ \text{tr}\ X'WY+\text{tr}\ Y'W^\star Y,
\end{equation}
with $W_\star$ and $W^\star$ the diagonal matrices with row-sums and column-sums of
$W$, as well as the constant
\begin{equation}
C:=\sum_{i=1}^n\sum_{k=1}^Kw_{ik}\delta_{ik}^2
\end{equation}
\end{subequations}

With these definitions
\begin{equation}
\sigma(X,Y)=C-2\ \rho(X,Y)+\eta^2(X,Y),
\end{equation}
and $\sigma(X,Y)\leq\sigma(X,Y;\tilde X, \tilde Y)$, where
\begin{equation}
\sigma(X,Y;\tilde X, \tilde Y):= C-2\ \rho(X,Y;\tilde X,\tilde Y)+\eta^2(X,Y),
\end{equation}
with equality if $X=\tilde X$ and $Y=\tilde Y$. In a majorization (or MM)
iteration we minimize $\sigma(X,Y;\tilde X, \tilde Y)$ over $X$ and $Y$. If the
minimum is attained at $\tilde X$ and $\tilde Y$ we stop the algorithm, because
we have found a stationary point. If not we have found updates $X^+$ and $Y^+$
such that
\begin{equation}
\sigma(X^+,Y^+)\leq\sigma(X^+,Y^+;\tilde X, \tilde Y)<
\sigma(\tilde X,\tilde Y;\tilde X, \tilde Y)=\sigma(\tilde X,\tilde Y).
\end{equation}

The stationary equations for minimizing $\sigma(X,Y;\tilde X, \tilde Y)$ are
\begin{subequations}
\begin{align}
X&=W_\star^{-1}(WY+B_\star(\tilde X,\tilde Y)\tilde X-B(\tilde X,\tilde Y)\tilde Y),\label{eq-xy1}\\
Y&=\{W^\star\}^{-1}(W'X+B^\star(\tilde X,\tilde Y)\tilde Y-B'(\tilde X, \tilde Y)\tilde X).\label{eq-xy2}
\end{align}
\end{subequations}
Alternatively
\begin{equation}
\begin{bmatrix}
W_\star&-W\\
-W'&W^\star
\end{bmatrix}
\begin{bmatrix}
X\\Y
\end{bmatrix}=
\begin{bmatrix}
B_\star(\tilde X,\tilde Y)&-B(\tilde X,\tilde Y)\\
-B'(\tilde X, \tilde Y)&B^\star(\tilde X,\tilde Y)
\end{bmatrix}
\begin{bmatrix}
\tilde X\\\tilde Y
\end{bmatrix}\label{eq-xy3}
\end{equation}

The next step in each iteration could be to directly solve the system
\eqref{eq-xy3} for $X$ and $Y$. This approach has the disadvantage that the
order is $n+m$, which can be quite large in applications. We also have to take
into account that the coefficient matrix of \eqref{eq-xy3} is singular.
Alternatively we can substitute the expression for $X$ from \eqref{eq-xy1} in
\eqref{eq-xy2} and then solve the resulting equation for $Y$. Or we can solve
the system in \eqref{eq-xy1} and \eqref{eq-xy2} iteratively by ALS. Choose a
$Y$, then compute $X$ from \eqref{eq-xy1}, then use that $X$ in \eqref{eq-xy2}
to find a new $Y$, and so on.

In this paper we use an approach similar to ALS, which could be called MALS for
Majorization by Alternating Least Squares. In the first step of a MALS cycle we
majorize $\sigma(X,\tilde Y)$ and minimize
$\sigma(X,\tilde Y;\tilde X,\tilde Y)$. Thus we treat the loss as a function of
$X$ only, and by the earlier results in this section we have
\begin{subequations}
\begin{equation}
X^+=W_\star^{-1}(W\tilde Y+B_\star(\tilde X,\tilde Y)\tilde X-B(\tilde X,\tilde Y)\tilde Y).\label{eq-xy4}
\end{equation}
The next step, in the same MALS iteration, is to us this new $X^+$ and minimize
$\sigma(X^+,Y;X^+,\tilde Y)$ over $Y$, giving
\begin{equation}
Y^+=\{W^\star\}^{-1}(W'X^+)+B^\star(X^+,\tilde Y)\tilde Y-B'(X^+, \tilde Y)X^+)
\end{equation}
\end{subequations}
Then proceed to the next two-step MALS iteration, using the new $X^+$ and $Y^+$.

One advantage of MALS, shared with ALS, is that the only matrix inversion needed
is of thee diagonal weight marginals. Also there is no need to work with square
symmetric matrices of order $n+m$. But the main advantage of MALS is that it can
easioly incorporate various constraints on $X$ and $Y$. Remember that
\begin{equation}
\sigma(X, \tilde Y;\tilde X,\tilde Y)=\text{tr}\ X'W_\star X-2\ \text{tr}\ X'\{W\tilde Y-\text{tr}\ X'B_\star(\tilde X,\tilde Y)\tilde X+\text{tr}\ X'B(\tilde X,\tilde Y)\tilde Y\}+C,
\end{equation}
where $C$ is now used generically for terms not involving $X$. Define
\begin{equation}
\overline X:=W\tilde Y-\text{tr}\ X'B_\star(\tilde X,\tilde Y)\tilde X+\text{tr}\ X'B(\tilde X,\tilde Y)\tilde Y
\end{equation}
Then, after completing the square,
\begin{subequations}
\begin{equation}
\sigma(X, \tilde Y;\tilde X,\tilde Y)=\text{tr}\ (X-\overline X)'W_\star(X-\overline X)+C.
\end{equation}
Thus minimizing $\sigma(X, \tilde Y;\tilde X,\tilde Y)$ over $X$ that must
satisfy some constraints means projecting $\overline X$ in the diagonal
$W_\star$ metric on the set defined by the constraints. Of course in the
same way
\begin{equation}
\sigma(\tilde X, Y;\tilde X,\tilde Y)=\text{tr}\ (Y-\overline Y)'W^\star(Y-\overline Y)+C.
\end{equation}
\end{subequations}

\sectionbreak

# Constraints

## Regression Constraints

In this paper we (optionally) use the constraints
\begin{equation}
\sum_{\ell=1}^{k_j}w_{i\ell}^j(\delta_{i\ell}^j-\overline{\delta}_{ij})^2=w_{i\star}^j,\label{eq-norm}
\end{equation}
for all $i$ and $j$, with $\smash{\overline{\delta}_{ij}}$ the weighted mean of
the $\smash{\delta_{i\ell}^j}$. In addition, of course, the $\delta_{i\ell}^j$
must satisfy the ordinal constraints. Constraint \eqref{eq-norm} explicitly
excludes the possibility that the $\smash{\delta_{i\ell}^j}$ for given $i$ and
$j$ are equal for all $\ell$.

Suppose $y$ is a vector with $n$ elements. The problem we study is to minimize
$\sigma(x):=\|x-y\|_W^2$ over $x\in K$ and $\|x-\overline x e\|_W^2$ is equal to
some constant $c$. Here

- $K$ is the cone of vectors defined by a number of inequalities of the form
  $x_i\leq x_j$,
- $e$ is the vector with all $n$ elements equal to one,
- $W$ is a diagonal matrix of weights,
- $\|x\|^2_W=x'Wx$ is the square of the $W$-norm of $x$,
- $\overline x$ is the weighted mean $e'Wx/e'We$.

Define $J:=I-ee'W/(e'We)$ and $\tilde x=Jx$. Then $Jx=x-\overline x e$. Then
$x'Jx=\tilde x'W\tilde x$ and $x\in K$ if and only if $\tilde x\in K$. Also
$\tilde x= Jx$ if and only if $x=\tilde x+\beta e$. So the transformed problem
is to minimize $\|\tilde x+\beta e-y\|_W$ over $\beta$ and $\tilde x\in K$, with
$e'W\tilde x=0$ and $\tilde x'W\tilde x=c$. The minimizer for $\beta$ is
$\overline{y}:=e'Wy/e'We$, and thus the equivalent "reduced" problem becomes
minimizing $\|\tilde x- Jy\|_W^2$ over $\tilde x\in K$, $\tilde x'W\tilde x=c$,
and $e'W\tilde x=0$.

Forget about the constraint $e'W\tilde x=0$ for now. If $M$ is the monotone
regression operator it follows that the solution for $\tilde x$ is
$\sqrt{c}M(Jy)/\|M(Jy)\|_W$, which automatically satisfies $e'W\tilde x=0$
(@bauschke_bui_wang_18). Thus the solution of our original problem is
$$
\hat x=\sqrt{c}\frac{M(y-\overline{y})}{\|M(y-\overline{y})\|_W}+\overline{y}e.
$$
This can be rewritten in a perhaps more revealing way. Now
$M(y-\overline{y}e)=M(y)-\overline{y}e$ and $e'WM(y)=e'Wy=e'We\overline{y}$.
Thus $\|M(y-\overline y)\|_W^2=\|M(y)\|_W^2-e'We\overline{M(y)}^2$ so that
$$
\hat x=\frac{M(y)-\overline{y}e}{\|M(y-\overline{y})\|_W}+\overline{y}e=
\frac{1}{\|M(y-\overline{y})\|_W}M(y)+\left\{1-\frac{1}{\|M(y-\overline{y})\|_W}\right\}\overline{y}e
$$

## Configuration Constraints

Separation

### Column-Centering

If we require $X'W_re=0$ then the stationary equations are
$W_r(X-\overline X)=W_re\lambda'$ and thus $X=\overline X+e\lambda'$ with
$\lambda'=-e'W_r\overline X/e'W_re$ or
$$
X=(I-\frac{ee'W_r}{e'W_re})\overline X
$$

This is different from updating $X$ as $J_n\tilde X$ and $Y$ as
$\tilde Y-n^{-1}ee'\tilde X$. With this alternative update, in which we subtract
the column-means of $\tilde X$ from both $\tilde X$ and $\tilde Y$, the
distances are the same before and after centering. Thus the distances in all
iterations, and consequently the loss function values, are the same with or
without this type of centering.

### Orthogonality

If we require $X'W_\star X=I$ we have the stationary equations $\overline X=XM$,
with $M$ a symmetric matrix of Lagrange multipliers. Thus
$$
X=\text{procrustus}(\overline X):=\overline X(\overline X'W_\star \overline X)^{-\frac12}.
$$
If we require both $X'W_\star X=I$ and $X'W_\star e=0$ we have
$$
X=\text{procrustus}((I-\frac{ee'W_\star }{e'W_\star e})\overline X)
$$

### Rank Constraints

One of the main themes in @gifi_B_90 is that the $k_j\times p$ matrices of
category-scores can be constrained to be of rank $r<p$. We will restrict
ourselves in this paper to $r=1$, which defines what Gifi calls *single
quantifications*. If all variables are restricted to have single quantifications
then a HGA becomes equivalent to a PCA of the transformed variables.

In VHA the Voronoi regions of single variables become parallel strips. The
category-points of a variable are on a line through the origin, and the Voronoi
lines or planes are parallel and perpendicular to the line with the
category-points.

$$
\text{tr}\ (Y-za')'W_c(Y-za')=-2z'W_cYa+z'W_cz.a'a
$$

So rank one approximation of $\tilde Y+n^{-1}ee'(X-\tilde X)$. If $X$ is
column-centered then of course $\tilde Y-ee'\tilde X)$.

### Centroid

A more complicated restriction is $Y=D^{-1}G'X$, possibly together with $X'X=I$
and $e'X=0$. Minimization over $X$ and the $Y_j$ is no longer separated. In
fact, the *centroid constraints* effectively eliminate the $Y_j$ from the
optimization problem. We have to adapt the majorization method to take these new
constraints into account.

If $Y_j=D_j^{-1}G_j'X$ then $d^j_{i\ll}(X,Y_j)$ only depends on $X$ and we have
\begin{subequations}
\begin{equation}
\{d_{i\ell}^j(X)\}^2=\text{tr}\ X'\left\{e_ie_i'-2\ \{d^j_\ell\}^{-1}g^j_\ell e_i'+\{d^j_\ell\}^{-2}g^j_\ell\{g^j_\ell\}'\right\}X.
\end{equation}
Moreover for each pair $X$ and $\tilde X$ we have, by Cauchy-=Schwartz,
\begin{equation}
d_{i\ell}^j(X)\geq\frac{1}{d_{i\ell}^j(\tilde X)}\text{tr}\ X'\{e_ie_i'-2\ \{d^j_\ell\}^{-1}g^j_\ell e_i'+\{d^j_\ell\}^{-2}g^j_\ell\{g^j_\ell\}'\}\tilde X
\end{equation}
\end{subequations}

Suppose $R_W^j$ and $C_W^j$ are diagonal matrices with row-sums and column=sums
of $W_j$. Then
\begin{subequations}
\begin{equation}
\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}^j\{d_{i\ell}^j(X)\}^2=
\text{tr}\ X'\left\{R_W^j-G_jD_j^{-1}W_j'-W_jD_j^{-1}G_j'+G_j(C_W^jD_j^{-2})G_j'\right\}X,
\end{equation}
and
\begin{equation}
\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}^jd_{i\ell}^j(X)\geq
\text{tr}\ X'\left\{R_{\tilde B}^j-G_jD_j^{-1}\tilde B_j'-\tilde B_j'D_j^{-1}G_j'+G_j(C_{\tilde B}^jD_j^{-2})G_j'\right\}\tilde X.\label{eq-ozob}
\end{equation}
\end{subequations}
In \eqref{eq-ozob} the matrix $\tilde B_j$ is defined as
\begin{equation}
\tilde b_{i\ell}^j:=\begin{cases}
w_{i\ell}^j\frac{\delta_{i\ell}^j}{d_{i\ell}^j(\tilde X)}&\text{ if }d_{i\ell}^j(\tilde X)>0,\\
0&\text{ if }d_{i\ell}^j(\tilde X)=0,
\end{cases}
\end{equation}
and $R_{\tilde B}^j$ and $C_{\tilde B}^j$ are the diagonal matrices of its row-sums
and column-sums.

\sectionbreak

# Software

## Main

The main program voronoiHomogeneityAnalysis(), aliased as vha(), has the
following arguments (with default values).

```{r}
#| label: arguments
#| eval: false

voronoiHomogeneityAnalysis <- function(theData,  # n x m matrix with categorical variables
                                       ndim = 2, # dimensionality
                                       wght = NULL,
                                       dnorm = FALSE,
                                       xcent = FALSE,
                                       xnorm = FALSE,
                                       yrank = ndim,
                                       itmax = 1000,
                                       eps = 1e-6,
                                       aps = 1e-6,
                                       verbose = TRUE)
```

- *theData* is an $n\times m$ matrix with $n$ observations on $m$categorical
  variables.
- *ndim* is the number of dimensions of the solution.
- *wght* is either NULL or a list of $n\times k_j$ matrices. If wght is NULL all
  weights are one.
- *dnorm* if TRUE the monotone regression is normalized.
- *xcent* if TRUE object scores are W-centered.
- *xnorm* if TRUE object scores are W-orthonormal.
- *yrank* rank restrictions on category-scores. If the length of yrank is one it
  is expanded to a vector of length $m$.
- *itmax* is the maximum number of iterations.
- *eps* is the cutoff for the change in distances from one iteration to another.
- *aps* is the cutoff for the loss function value. If loss is smaller than *aps*
  the program stops.
- *verbose* if TRUE iteration information is printed to stdout.

The value returned by the program is a list.

```{r}
#| label: output
#| eval: false

  return(
    list(
      xmat = xnew,
      ymat = ynew,
      dmat = dnew,
      dhat = dhat,
      xini = xini,
      yini = yini,
      itel = itel,
      loss = snew,
      indi = indi,
      ncat = ncat,
      wght = wght
    )
  )
```

- *xmat* object scores, an $n\times p$ matrix, with $p$ equal to *ndim*.
- *ymat* category-scores, a list of *m* matrices of dimension $k_j\times p$.
- *dmat* distances, a list of *m* matrices of dimension $n\times k_j$.
- *dhat* disparities, a list of *m* matrices of dimension $n\times k_j$.
- *xini* the HGA solution for the object scores,
- *yini* the HGA solution for the category-scores,
- *itel* number of iterations.
- *loss* final value of stress.
- *indi* indicators, a list of *m* matrices of dimension $n\times k_j$.
- *ncat* number of categories of the $m$ variables.
- *wght* weights, a list of *m* matrices of dimension $n\times k_j$.

## Subroutines

## Plots

\sectionbreak

# Examples

## Small Example

We start with the small example we used before to illustrate HGA.

```{r}
#| label: smallvha
#| fig-align: "center"
#| echo: false

h_small <- vha(small, itmax = 1000, eps = 1e-10, verbose = FALSE)
par(mfrow = c(1, 3))
plotVHA(h_small$xmat, h_small$ymat[[1]], h_small$indi[[1]], ncat_small[1], xlab = small[, 1], ylab = labs_small[[1]], pdf = FALSE)
plotVHA(h_small$xmat, h_small$ymat[[2]], h_small$indi[[2]], h_small$ncat[2], xlab = small[, 2], ylab = labs_small[[2]], pdf = FALSE)
plotVHA(h_small$xmat, h_small$ymat[[3]], h_small$indi[[3]], h_small$ncat[3], xlab = small[, 3], ylab = labs_small[[3]], pdf = FALSE)
```
Loss is `r formatC(h_small$loss, digits = 10, format = "f")` after
`r h_small$itel` iterations. If we compare the plots from the Voronoi analysis
with the star plots from the initial HGA for this example we see that the
algorithm did not have to change much to obtain perfect fit. One thing that does
change is that the iterations moves the category-points "p" and "r" for variable
2 very close to each other. At the point where we stop the iterations they are
still distinguishable, and thus their perpendicular bisector is still a line in
the Voronoi plot. If we would continue iterating "p" and "r" would coincide and
there would only be two Voronoi regions: one containing category-point "q" and
object-point 2, and one containing all other points. one containing all other.

```{r}
#| label: redo
#| echo: false
#| fig-align: "center"

plotVHA(h_small$xmat, h_small$ymat[[2]][1:2, ], h_small$indi[[2]], 23, xlab = small[, 2], ylab = c("p r", "q"), pdf = FALSE)
```

## GALO

Our next example use the GALO data from @peschar_75. In the github voronoi
repository the data are in data/galo.R. Subjects are 1290 children who in
1959/1960 were in sixth grade of primary school in the city of Groningen.
Variables are SEX, IQ (in nine ordered categories), ADV (sixth grade teachers
advice on the form of secondary education appropriate for the student), and SES.
The categories of ADV are as follows (alphabetically).

- *Agr* Agricultural.
- *Ext* Extended primary (like middle school).
- *Gen* General (like high school).
- *Grls* Extended primary for girls (like middle school).
- *Man* Vocational, including housekeeping.
- *None* No secondary education.

The categories of SES are as follows (again alphabetically).

- *LowWC* Lower white collar.
- *MidWC* Middle white collar.
- *Prof* Professional.
- *Shop* Shopkeepers.
- *Skil* Skilled manual.
- *Unsk* Unskilled Manual.

We start with an HGA of the galo data. The initial value of the loss function is
0.3158339145. The loss function value is computed by using
\begin{equation}
\sigma_\star(X,Y_1,\cdots,Y_m):=\min_{\Delta_1,\cdots,\Delta_m}\sigma(X,Y_1,\cdots,Y_m,\Delta_1,\cdots,\Delta_m),\label{eq-losshga}
\end{equation}
where $X$ and the $Y_j$ are the HGA solution, and where the minimum over the
$\Delta_j$ is computed by requiring that
$\smash{\delta_{i\nu}^j\leq\delta_{i\ell}^j}$ if $\smash{g_{i\nu}^j=1}$. This is
*not* the HGA loss function \eqref{eq-loss2}, in which the minimum over the
$\Delta_j$ is computed by requiring that $\delta_{i\ell}^j=0$ whenever
$g_{i\ell}^j=1$. Thus we evaluate the HGA solution using the VGA loss function.

For any HGA or VGA solution we can compute a Shepard diagram by plotting the
distances and disparities, pooled over all objects (individuals) and variables.
The weak regression requirements imply that many disparities will be equal to
the corresponding distances. This is illustrated by the diagonal of
@fig-galo_d_dhat_hga.

```{r, verbose=FALSE}
#| label: galovha
#| echo: false
#| cache: true

hgalo <- vha(galo[, 1:4], itmax = 1000, verbose = FALSE)
```

```{r}
#| label: galo_bnds_hga
#| echo: false

gmax <- max(c(max(hgalo$xini), sapply(hgalo$yini, max)))
gmin <- min(c(min(hgalo$xini), sapply(hgalo$yini, min)))
```

```{r}
#| label: fig-galo_d_dhat_hga
#| fig-align: "center"
#| echo: false
#| fig-width: 10
#| fig-height: 10
#| fig-cap: "Galo D-Dhat Plot HGA"

dold <- rep(list(0), 4)
dhat <- rep(list(0), 4)
for (j in 1:4) {
    dold[[j]] <- makeDmat(hgalo$xini, hgalo$yini[[j]])
    dhat[[j]] <- monotone(dold[[j]], hgalo$ncat[j], hgalo$indi[[j]])
  }
dmatmat <- makeMatrixFromList(dold)
dhatmat <- makeMatrixFromList(dhat)
par(pty = "s")
dmatmat <- makeMatrixFromList(dold)
dhatmat <- makeMatrixFromList(dhat)
plot(dmatmat, dhatmat, col = "RED", pch = 16, xlab = "Distances", ylab = "Disparities")
```

The HGA solution can also be used to produce Voronoi plots, which are given in
@fig-galo_IQ_HGA for IQ. We see the familiar horseshoe shape of the object score
cloud (the Guttman effect, see @deleeuw_C_23). The category-scores are largely
in order along the horseshoe, and as a consequence the Voronoi regions are
unbounded and emanate from a point inside the horseshoe. There are some
relatively monor deviations from that pattern. Specifically category 1 does not
fit the pattern, but that is not too serious because there are only five
individuals in that category. There are quite a few misclassifications, i.e.
objects in the "wrong" Voronoi region.

```{r}
#| label: fig-galo_IQ_HGA
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 10
#| echo: false
#| fig-cap: "Galo IQ HGA"

par(pty = "s")
plot(voronoi(hgalo$yini[[2]], ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(hgalo$yini[[2]], as.character(1:9), col = "RED", cex = 1.5)
text(hgalo$xini, as.character(galo[, 2]), col = "BLUE")
```

```{r}
#| label: galo_bnds
#| echo: false

gmax <- max(c(max(hgalo$xmat), sapply(hgalo$ymat, max)))
gmin <- min(c(min(hgalo$xmat), sapply(hgalo$ymat, min)))
```

The VHA solution has loss equal to
`r formatC(hgalo$loss, digits = 10, format = "f")` after `r hgalo$itel`
iterations. The voronoi function stopped because the maximum number of
iterations was reached. The $d-\hat d$ plot in @fig-galo_d_dhat shows a
near-perfect fit.

```{r}
#| label: fig-galo_d_dhat
#| fig-align: "center"
#| echo: false
#| fig-width: 10
#| fig-height: 10
#| fig-cap: "Galo D-Dhat Plot VHA"

par(pty = "s")
dmatmat <- makeMatrixFromList(hgalo$dmat)
dhatmat <- makeMatrixFromList(hgalo$dhat)
plot(dmatmat, dhatmat, col = "RED", pch = 16, xlab = "Distances", ylab = "Disparities")
```

```{r}
#| label: fig-galo_IQ
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 10
#| echo: false
#| fig-cap: "Galo IQ VHA"

par(pty = "s")
plot(voronoi(hgalo$ymat[[2]], ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(hgalo$ymat[[2]], as.character(1:9), col = "RED", cex = 1.5)
text(hgalo$xmat, as.character(galo[, 2]), col = "BLUE")
```

@fig-galo_IQ shows some characteristic properties of Voronoi plots. If the
objects in two categories are not well separated the object-points will be on
the perpendicular bisector of the category pair. This is partial degeneracy due
to requiring weak monotonicity. The algorithm cannot impose strict monotonicity
by placing an object in the interior of its Voronoi region, and compromises by
putting it on the boundary. Thus the VHA plot will show object-points on lines.
Some category scores are very close, which makes the Voronoi diagram unstable.
The solution for the category-scores for IQ is

```{r}
#| label: galo_y2
#| echo: false

y <- hgalo$ymat[[2]]
rownames(y) <- lab2
y <- cbind(y, colSums(hgalo$indi[[2]]))
colnames(y) <- c("dim1", "dim2", "freq")
print(y)
```

We see that categories 2 and 3 map into basically the same point. The same is
true for categories 4 and 5, and for categories 6 and 7. So instead of nine
points there are really only six. We can clean up the Voronoi diagram by using
only those six categories. Note that we do not recode the data and redo the
analysis, we just use the same plot with the same object-scores and just change
the diagram. The result is in @fig-galo_IQ2.

```{r}
#| label: fig-galo_IQ2
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 10
#| echo: false
#| fig-cap: "Galo IQ Corrected VHA"

par(pty = "s")
ycor <- hgalo$ymat[[2]][c(1, 2, 4, 6, 8, 9), ]
plot(voronoi(ycor, ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(ycor, c("1", "2 3", "4 5", "6 7", "8", "9"), col = "RED", cex = 1.5)
text(hgalo$xmat, as.character(galo[, 2]), col = "BLUE", cex = .75)
```

## Cetacea

The data in @vescia_85 describe 36 whales, porpoises and dolphins using 15
categorical variables. In the github voronoi repository the data are in
data/cetacea.R. Appendix C gives the common names of the 36 species and the
names of the variables.

The variables are mostly physical characteristics, but they also include habitat
and feeding. The average number of categories per variable is about four, and
almost all variables are "nominal" (i.e. not "ordinal" and not "numerical"). In
the same book @vanderburg_85 presents a secondary analysis of the data with HGA
and @guttman_85 applies his MSA-I technique (which uses HGA as an initial
configuration for its iterations).

We start with an HGA of the cetacea data. The initial value of the loss function
... is 0.9945577132.

```{r, verbose=FALSE}
#| label: cetaceavha
#| echo: false
#| cache: true

hcetacea <- vha(cetacea, itmax = 1000, verbose = FALSE)
```

```{r}
#| label: cetacea_bnds_hga
#| echo: false

gmax <- max(c(max(hcetacea$xini), sapply(hcetacea$yini, max)))
gmin <- min(c(min(hcetacea$xini), sapply(hcetacea$yini, min)))
```

```{r}
#| label: cetacea_d_dhat_hga
#| fig-align: "center"
#| echo: false
#| fig-width: 10
#| fig-height: 10
#| fig-cap: "Cetacea D-Dhat Plot HGA"

dold <- rep(list(0), 15)
dhat <- rep(list(0), 15)
for (j in 1:15) {
    dold[[j]] <- makeDmat(hcetacea$xini, hcetacea$yini[[j]])
    dhat[[j]] <- monotone(dold[[j]], hcetacea$ncat[j], hcetacea$indi[[j]])
  }
dmatmat <- makeMatrixFromList(dold)
dhatmat <- makeMatrixFromList(dhat)
par(pty = "s")
dmatmat <- makeMatrixFromList(dold)
dhatmat <- makeMatrixFromList(dhat)
plot(dmatmat, dhatmat, col = "RED", pch = 16, xlab = "Distances", ylab = "Disparities")
```

```{r}
#| label: cetacea_xmat_HGA
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 10
#| echo: false
#| fig-cap: "Cetacea Object Scores HGA"

par(pty = "s")
plot(hcetacea$xini, xlim = 1.1 * c(gmin, gmax), ylim = 1.1 * c(gmin, gmax), xlab = "dim 1", ylab = "dim 2", type = "n")
text(hcetacea$xini, rownames(cetacea), col = "RED", cex = 1.00)
```

```{r}
#| label: cetacea_flippers_HGA
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 10
#| echo: false
#| fig-cap: "Cetacea Flippers HGA"

par(pty = "s")
plot(voronoi(hcetacea$yini[[6]], ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(hcetacea$yini[[6]], cetacea_cats[[6]], col = "RED", cex = 1.5)
text(hcetacea$xini, cetacea_cats[[6]][hcetacea$indi[[6]] %*% 1:4], col = "BLUE", cex = .75)
```

```{r}
#| label: cetacea_bnds
#| echo: false

gmax <- max(c(max(hcetacea$xmat), sapply(hcetacea$ymat, max)))
gmin <- min(c(min(hcetacea$xmat), sapply(hcetacea$ymat, min)))
```

Loss is `r formatC(hcetacea$loss, digits = 10, format = "f")` after
`r hcetacea$itel` iterations.

```{r}
#| label: cetacea_d_dhat
#| fig-align: "center"
#| echo: false
#| fig-width: 10
#| fig-height: 10
#| fig-cap: "Cetacea D-Dhat Plot VHA"

par(pty = "s")
dmatmat <- makeMatrixFromList(hcetacea$dmat)
dhatmat <- makeMatrixFromList(hcetacea$dhat)
plot(dmatmat, dhatmat, col = "RED", pch = 16, xlab = "Distances", ylab = "Disparities")
```

```{r}
#| label: cetacea_y6
#| echo: false

y <- hcetacea$ymat[[6]]
rownames(y) <- cetacea_cats[[6]]
y <- cbind(y, colSums(hcetacea$indi[[6]]))
colnames(y) <- c("dim1", "dim2", "freq")
print(y)
```

```{r}
#| label: cetcea_flippers
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 10
#| echo: false
#| fig-cap: "Cetacea Flippers"

par(pty = "s")
plot(voronoi(hcetacea$ymat[[6]], ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(hcetacea$ymat[[6]], cetacea_cats[[6]], col = "RED", cex = 1.5)
text(hcetacea$xmat, cetacea_cats[[6]][hcetacea$indi[[6]]%*%1:4], col = "BLUE", cex = .75)
```

```{r, verbose=FALSE}
#| label: cetaceavhacc
#| echo: false
#| cache: true

hcetaceacc <- vha(cetacea, itmax = 1000, verbose = FALSE, xcent = TRUE, xnorm = TRUE)
```

Loss is `r formatC(hcetaceacc$loss, digits = 10, format = "f")` after
`r hcetaceacc$itel` iterations.

```{r}
#| label: cetacea_d_dhat_tt
#| fig-align: "center"
#| echo: false
#| fig-width: 10
#| fig-height: 10
#| fig-cap: "Cetacea D-Dhat Plot VHA XX"

par(pty = "s")
dmatmat <- makeMatrixFromList(hcetaceacc$dmat)
dhatmat <- makeMatrixFromList(hcetaceacc$dhat)
plot(dmatmat, dhatmat, col = "RED", pch = 16, xlab = "Distances", ylab = "Disparities")
```

```{r}
#| label: cetacea_bnds_tt
#| echo: false

gmax <- max(c(max(hcetaceacc$xmat), sapply(hcetaceacc$ymat, max)))
gmin <- min(c(min(hcetaceacc$xmat), sapply(hcetaceacc$ymat, min)))
```

```{r}
#| label: cetcea_flippers_tt
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 10
#| echo: false
#| fig-cap: "Cetacea Flippers VHA XX"

par(pty = "s")
plot(voronoi(hcetaceacc$ymat[[6]], ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(hcetaceacc$ymat[[6]], cetacea_cats[[6]], col = "RED", cex = 1.5)
text(hcetaceacc$xmat, cetacea_cats[[6]][hcetaceacc$indi[[6]]%*%1:4], col = "BLUE", cex = .75)
```
\sectionbreak

# Appendix: Cetacea

```{r}
#| label: cetacea_rownames
#| echo: false

r <- noquote(rownames(cetacea))
print(r, right = TRUE)
```

```{r}
#| label: cetacea_colnames
#| echo: false

r <- noquote(colnames(cetacea))
print(r, right = TRUE)
```


\sectionbreak

# Appendix: Binary Monotone Regression

Suppose $y$ and $w\geq 0$ are given vectors with $n$ elements. Suppose
$N_1$ and $N_0$ are a partition of $N:=\{1,2,\cdots n\}$. The Binary
Monotone Regression Problem is to find $x$ with $x_i\leq x_k$ for all 
$i\in N_0$ and $k\in N_1$
such that
$$
\sigma(x)=\frac12\sum_{i\in N} w_i(x_i-y_i)^2
$$
is minimized. Equivalently we want to find $\alpha$ such that
$x_i\leq\alpha$ for $i\in N_0$ and $x_i\geq\alpha$ for
$i\in N_1$.
Thus loss can be rewritten as a function of the single variable $\alpha$.
$$
\sigma(\alpha)=\frac12\sum_{i\in N_0} w_i\min(\alpha-y_i,0)^2+\frac12\sum_{i\in N_1} w_i\max(\alpha-y_i,0)^2.
$$
This function is piecewise quadratic, convex, and differentiable (although not twice
differentiable). Its derivative, which is increasing, is
$$
\mathcal{D}\sigma(\alpha)=\sum_{i\in N_0} w_i(\min(\alpha-y_i,0)+\sum_{i\in N_1} w_i(\max(\alpha-y_i, 0)
$$
The function binaryMonotoneRegression(), with the alias bmr(), minimizes $\sigma$ over
$\alpha$. We sort $y$, set $\alpha$ equal to the smallest $y_i$, and compute 
$\mathcal{D}\sigma(\alpha)$. If the derivative is zero, we are done. If it is
negative we set $\alpha$ to the next smallest $y_i$. We continue this until we arrive
at an $\alpha$ for which the derivative is either zero or positive. If it is
zero we stop. If it is positive then the minimum is between this $\alpha$ and
the previous $\alpha$, for which the derivative was still negative. Since the derivative is linear on the interval between the two values of $\alpha$ we can use linear interpolation to find the value for which it is zero.

# References
