---
title: Voronoi Homogeneity Analysis of Categorical Data
author: 
    - name: Jan de Leeuw
      orcid: 0000-0003-1420-1797
      email: jan@deleeuwpdx.net
      affiliation: 
        - name: University of California Los Angeles
          city: Los Angeles
          state: CA
          url: www.ucla.edu
      license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
   pdf:
    fontsize: 12pt
    include-in-header: 
     - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
   html:
    fontsize: 12pt
    include-in-header: 
     - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: TBD
---

```{r loadpackages, echo = FALSE}
suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
suppressPackageStartupMessages(library(MASS, quietly = TRUE))
suppressPackageStartupMessages(library(RSpectra, quietly = TRUE))
suppressPackageStartupMessages(library(deldir, quietly = TRUE))
suppressPackageStartupMessages(library(dismo, quietly = TRUE))
```

```{r sourcecode, echo = FALSE}
source("small.R")
source("galo.R")
source("mca.R")
source("majorization.R")
source("auxiliaries.R")
source("voronoi.R")
source("plotVHA.R")
```

\sectionbreak 

**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw/voronoi> 

\sectionbreak

# Introduction

## Homogeneity Analysis

*Homogeneity Analysis (HGA)*, a.k.a. *Multiple Correspondence Analysis (MCA)*, can be introduced and implemented in many different ways\footnote{Since PCA and MCA have three-letter
acronyms I added the G of Gifi to the HA of Homogeneity Analysis.}. We refer to
@guttman_41, @tenenhaus_young_85, @bekker_deleeuw_C_88, @gifi_B_90, @greenacre_blasius_06, @leroux_rouanet_10, @husson_josse_saporta_16, and @deleeuw_C_23 for a bouquet of
discussions and interpretations of the equations and algorithms
defining the HGA technique.

A *variable* is a function that maps on a (finite or infinite) domain of *objects*
into a (finite or infinite) codomain of *values*.
In HGA the data are the values of $m$ variables on the same *carrier*, a 
finite subset the domain consisting of $n$ objects. 
Each variable $j$ has a finite number $k_j$ of *categories*.  The categories
constitute the *range* of the variable for the $n$ objects in the carrier. Note that all
variables have the same carrier. Each object maps into a unique category, and each category 
contains a subset of objects from the carrier. We also say that a category has a number of objects that "belong to" the category.

Here is a small example with $n=10$ and $m=3$, taken from @gifi_B_90, page 65.
```{r small, echo = FALSE}
print(small)
```

The key mathematical object in Homogeneity Analysis is the
*indicator matrix*, which is a classical way to code categorical data\footnote{Since the indicator matrix is the core of the Gifi 
system we avoid the irreverent alternative "dummy coding".}.
For a variable $j$ with $k_j$ categories it is an $n\times k_j$ binary
matrix $G_j$ in which each row $i$ contains a single element equal to one,
indicating into which category the object $i$ maps for variable $j$.
The cross product $D_j:=G_j'G_j$ is diagonal, with the *marginal
frequencies* of the categories on the diagonal. For two different variables the crossproduct $C_{j\ell}=G_j'G_\ell$ is the $k_j\times k_\ell$ *cross table* or *contingency table* of the two variables. Here are the three 
indicator matrices of our small example, next to each other.

```{r indi, echo = FALSE}
g1 <- makeIndicator(small[, 1])
g2 <- makeIndicator(small[, 2])
g3 <- makeIndicator(small[, 3])
d1 <- diag(1 / colSums(g1))
d2 <- diag(1 / colSums(g2))
d3 <- diag(1 / colSums(g3))
gg <- noquote(cbind("||", g1, "||", g2, "||", g3, "||"))
colnames(gg) <- c("", "a", "b", "c", "", "p", "q", "r", "", "u", "v", "")
print(gg)
```

We now deviate from the usual Gifi and French approaches (@husson_josse_saporta_16) and define Homogeneity Analysis as a form of non-metric multidimensional scaling, using the *stress* loss function (@kruskal_64a, @kruskal_64b). Define\footnote{The symbol $:=$ is used for definitions}
\begin{equation}
\sigma(X,Y_1,\cdots,Y_m,\Delta_1,\cdots,\Delta_m) ):=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}^j(\delta_{i\ell}^j-d(x_i,y_\ell^j))^2,\label{eq-loss1}
\end{equation}
with $d(.,.)$ Euclidean distance.

Homogeneity analysis minimizes this loss over all $n\times p$ configurations $X$ satisfying $X'X=I$, over all $Y_j$, and over all $\Delta_j$ that satisfy $\delta_{i\ell}^j=0$ whenever $g_{i\ell}^j=1$. All other elements of the $\Delta_j$ are only constrained to be non-negative. 

If
\begin{equation}
\sigma_\star(X,Y_1,\cdots,Y_m):=\min_{\Delta_1,\cdots,\Delta_m}\sigma(X,Y_1,\cdots,Y_m,\Delta_1,\cdots,\Delta_m),\label{eq-loss2}
\end{equation}
then
\begin{equation}
\sigma_\star(X,Y_1,\cdots,Y_m)=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}g_{i\ell}^jd^2(x_i,y_\ell^j).\label{eq-loss3}
\end{equation}
Next define 
\begin{equation}
\sigma_{\star\star}(X):=\min_{Y_1,\cdots,Y_m}\sigma_\star(X;Y_1,\cdots,Y_m).\label{eq-loss4}
\end{equation}
The minimum of \eqref{eq-loss3} over the $Y_j$ is attained at the centroids $\hat Y_j:=D_j^{-1}G_j'X$, and if we substitute these optimal $\smash{\hat Y_j}$
in \eqref{eq-loss3} we see that $\sigma_{\star\star}(X)$ is indeed equal to the sum of squares of all lines defining the stars. Minimizing $\sigma_{\star\star}(X)$ over $X'X=I$, and thus minimizing \eqref{eq-loss1} over $(X, Y_j,\Delta_j)$, defines Homogeneity Analysis. 
Actually computing the solution turns out to be a straightforward eigenvalue-eigenvector problem. The details are in the publications we referenced at the beginning of this section.

A HGA in two dimensions produces of our small example the following three star plots. Objects
are numbered and in BLUE, categories have bigger labels and are in RED. 
Note that there are ten objects, but the sets of objects $(1,4,7)$, $(8,10)$, and $(6,9)$
have the same *profile* and consequently get mapped into the same object-point.
Thus there are at most six different object-points. The analysis minimizes the sum of squares of the length of all black lines in the three plots, under the condition that the matrix of object scores is column-centered and orthonormal. We can think of the results as a form
of *graph drawing*, where the graph is the union of the $m$ bipartite graphs of all $m$ "belonging to" relations (@michailidis_deleeuw_A_01).

```{r mca, echo = FALSE, fig.align = "center"}
haux <- mca(cbind(g1, g2, g3), 2)
y1 <- haux$ymat[1:3, ]
y2 <- haux$ymat[4:6, ]
y3 <- haux$ymat[7:8, ]
x <- haux$xmat
par(pty = "s")
par(mfrow = c(1, 3))
plot(rbind(x, y1), type = "n", xlab = "", ylab = "")
text(x, as.character(1:10), col = "BLUE", cex = .8)
text(y1, c("a", "b", "c"), col = "RED", cex = 1.5)
for (i in 1:10) {
  if (small[i, 1] == "a") lines(rbind(x[i,], y1[1, ]))
  if (small[i, 1] == "b") lines(rbind(x[i,], y1[2, ]))
  if (small[i, 1] == "c") lines(rbind(x[i,], y1[3, ]))
}
plot(rbind(x, y2), type = "n", xlab = "", ylab = "")
text(x, as.character(1:10), col = "BLUE", cex = .8)
text(y2, c("p", "q", "r"), col = "RED", cex = 1.5)
for (i in 1:10) {
  if (small[i, 2] == "p") lines(rbind(x[i,], y2[1, ]))
  if (small[i, 2] == "q") lines(rbind(x[i,], y2[2, ]))
  if (small[i, 2] == "r") lines(rbind(x[i,], y2[3, ]))
}
plot(rbind(x, y3), type = "n", xlab = "", ylab = "")
text(x, as.character(1:10), col = "BLUE", cex = .8)
text(y3, c("u", "v"), col = "RED", cex = 1.5)
for (i in 1:10) {
  if (small[i, 3] == "u") lines(rbind(x[i,], y3[1, ]))
  if (small[i, 3] == "v") lines(rbind(x[i,], y3[2, ]))
}
```


## Voronoi Analysis

Although using loss function \eqref{eq-loss1} makes HGA a form of non-metric multidimensional scaling there are some important differences with the more familiar applications of minimizing stress. There are no local minima in HGA, because it turns out to be a symmetric eigenvalue problem. Also, and perhaps most importantly, we do not
really expect the minimum loss function value to be zero, or even close
to zero. The loss function measures deviations from a very unrealistic
"model". Zero loss would imply all object-points coincide with 
all category-points they belong to. This
can only be realized, except in trivial cases, if all object and category-points map into a single point, which is prohibited by the orthonormality constraint on the object scores. Thus we get a perfectly determinate and stable solution, at low computational cost, but by imposing unrealistic constraints and a rather arbitrary normalization. We can get around this 
seemingly unfortunate situation by admitting we are not fitting or
testing a "model", but we are creating "best" drawings of the bipartite graphs 
that constitute the data.

The lack of a realistic model makes it interesting to generalize loss function \eqref{eq-loss1} to
\begin{equation}
\sigma(X,Y_1,\cdots,Y_m,\Delta_1,...,\Delta_m):=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}(\delta_{i\ell}^j-d(x_i,y_\ell^j))^2,\label{eq-loss5}
\end{equation}
with the constraints that
if $g_{i\ell}^j=1$ then merely $\delta_{i\ell}^j\leq\delta_{i\nu}^j$ for all
$\nu=1,\cdots,k_j$. This is significantly weaker than the homogeneity
analysis constraints, which require the $mn$ disparities $\smash{\delta_{i\ell}^j}$ to be zero.
If object $i$ belongs to category $\ell$ of 
variable $j$ we do no longer require the object-point to coincide
with the category-point, but we merely require that the object-point
is at least as close to category-point $\ell$ as to any of the
other $k_j-1$ category-points of that variable.

As a consequence, the plots corresponding with loss function \eqref{eq-loss5} are also different. As in Homogeneity Analysis there is a plot for each variable $j$ with both $n$ object-points and $k_j$ category-points. The $k_j$
category-points define $k_j$ *voronoi regions* that partition the
space. Voronoi region $\smash{\mathcal{V}^j_\ell}$ is the region of space
where the points are at least as close to category-point $\smash{y_\ell^j}$ as to any of the other category-points of variable $j$. And loss \eqref{eq-loss4} is zero (for a variable) if all object-points are in the "correct"
Voronoi region, i.e. in the Voronoi region of the category
point corresponding with the category they belong to. 

One (very inefficient, but conceptually easy) way of constructing
Voronoi regions in the plane is to draw the perpendicular bisectors of 
the (hypothetical) lines connecting all pairs of category-points of a variable. In higher dimensions $p>2$ the bisectors are planes of dimension $p-1$. The convex polyhedral regions thus formed, some bounded and some 
unbounded, define the Voronoi regions.
The literature on properties and computation on Voronoi regions 
is large, and we refer to the textbooks of @okabe_boots_sugihara_chiu_00
and @aurenhammer_klein_lee_13.

Since the category-point itself is obviously in its own Voronoi
region and since Voronoi regions are convex it follows that
if we make stars by connecting category-points with object
points as in Homogeneity Analysis then we actually require
each star to be in its Voronoi region. It is no longer 
true that a category-point is the centroid of points of the objects belonging to the category. The category-point can even be outside the 
convex hull of "its" object-points.

Voronoi Homogeneity Analysis of categorical data, as defined here, minimizes
loss function \eqref{eq-loss5} over $X$, over the $Y_j$ and
over the $\Delta_j$ satisfying the ordinal constraints. It is consequently
another form of non-metric (ordinal) multidimensional scaling. To be 
more precise, we have $m$ linked non-metric multidimensional unfolding
problems, which are linked because they share the same $X$. The
data are binary, and we use Kruskal's primary approach to ties
(@deleeuw_A_77).

The constraint $X'X=I$ from Homogeneity Analysis is no longer
needed, but some form of normalization is still necessary.
We still need to exclude the trivial solution in which all
points and regions collapse into a single point and all
$\Delta_j$ are zero. But, more urgently, we have to deal
with the trivial solutions familiar from multidimensional
unfolding (@heiser_81, @deleeuw_R_83, @busing_10). For example,
we can collapse all category-points into a single point and 
put all object-points on a sphere with the category-point
at its center. Then all distances between category and
object-points are the same and we can choose all dissimilarities
equal to that distance value and obtain stress zero. It does not
help to require $X'X=I$, because we can easily 
construct orthogonal configurations with all points on the 
sphere. Stronger normalization constraints are required.



# Algorithm

We use an *Alternating Least Squares* or *ALS* algorithm, similar to many
other non-metric muldimensional scaling methods, to minimize
loss \eqref{eq-loss5}. For context, see @borg_groenen_05.
We alternate minimization over $\Delta_1,\cdots,\Delta_m$
for given $X$ and $Y_1\cdots Y_m$ with minimization over $X$ and $Y_1\cdots Y_m$ for
given $\Delta_1,\cdots,\Delta_m$. Using superscript $(\mu)$ for the iteration index we can write for an ALS iteration
\begin{subequations}
\begin{align}
(\Delta_1^{(\mu)},\cdots,\Delta_m^{(\mu)})&=\mathop{\text{argmin}}_{(\Delta_1,\cdots,\Delta_m)}\sigma(X^{(\mu)},Y_1^{(\mu)},\cdots,Y_m^{(\mu)},\Delta_1,\cdots,\Delta_m),\label{eq-als1}\\
(X^{(\mu+1)},Y_1^{(\mu+1)},\cdots,Y_m^{(\mu+1)})&=\mathop{\text{argmin}}_{X,Y_1,\cdots,Y_m}\sigma(X,Y_1,\cdots,Y_m,\Delta_1^{(\mu)},\cdots,\Delta_m^{(\mu)}).\label{eq-als2}
\end{align}
\end{subequations}
The first substep \eqref{eq-als1} of iteration $(\mu)$ solves
a monotone regression problem for a simple partial order, in which
one of $k_j$ elements must be smaller than the other $k_j-1$ elements.
The second substep \eqref{eq-als2} solves a metric multidimensional scaling problem.
This subproblem cannot be solved completely, because it would need an 
infinite iterative process by itself. Thus the second substep \eqref{eq-als2}
consists of a finite number of *inner iterations* within the
second substep of a single ALS *outer iteration*. 

## Monotone Regression

The $nm$ monotone regressions are applied for each object $i$ and
each variable $j$. Forgetting about all these indices for a 
moment we need to minimize the sum of squares
$\|\delta-d\|^2$ over $\delta$, where one of the elements, 
say $\delta_\nu$, must be smaller than or equal to all others.

This can be solved by using monotone regression with the 
primary approach to ties
(@kruskal_64b, @deleeuw_A_77), treating all $\delta_\ell$
with $\ell\neq\nu$ as ties.

## Majorization

For the inner iterations we use standard *smacof* majorization steps
(@deleeuw_C_77). We define some of the entities normally used
in *smacof*. First we define the partitioned matrix\footnote{The $e_i$ are unit vectors with element
$i$ equal to one and the other elements zero. The number of elements of the $e_i$ follows from the
context.}
\begin{equation}
A_{ij}:=\begin{bmatrix}e_ie_i'&-e_ie_j'\\-e_je_i'&e_je_j'\end{bmatrix}.\label{eq-adef}
\end{equation}
Next
\begin{equation}
Z:=\begin{bmatrix}
X_{\ }\\
Y_1\\
\vdots\\
Y_m
\end{bmatrix}.\label{eq-zdef}
\end{equation}
With these definitions
\begin{equation}
d_{ij}(X,Y)=\sqrt{\text{tr}\ Z'A_{ij}Z},\label{eq-dz}
\end{equation}
and thus loss \eqref{eq-loss5} becomes
\begin{equation}
\sigma(X,Z):=\sum_{i=1}^n\sum_{\ell=1}^{k_j}(\delta_{i\ell}^j-\text{tr}\ Z'A_{i\ell}^jZ)^2,
\end{equation}



\begin{equation}
V:=\sum_{i=1}^n\sum_{j=1}^m A_{ij}=\begin{bmatrix}\hfill mI&-ee'\\-ee'&\hfill nI\end{bmatrix}\label{eq-vdef}
\end{equation}

If we define
\begin{align}
V_1&:=(n+m)^{-1}\begin{bmatrix}\frac{m}{n}E&-E\\-E&\frac{n}{m}E\end{bmatrix},\\
V_2&:=\begin{bmatrix}J_n&0\\0&0\end{bmatrix},\\
V_3&:=\begin{bmatrix}0&0\\0&J_m\end{bmatrix},
\end{align}
then
$$
V=(n+m)V_1+mV_2+nV_3
$$
Matrices $V_1,V_2,$ and $V_3$ are symmetric projectors on orthogonal
subspaces. It follows that $V^+$, the Moore-Penrose inverse3 of $V$,
is
$$
V^+=\frac{1}{n+m}V_1+\frac{1}{m}V_2+\frac{1}{n}V_3
$$
$$
B(X,Y):=\begin{bmatrix}B_{11}(X,Y)&B_{12}(X,Y)\\B_{21}(X,Y)&B_{22}(X,Y)\end{bmatrix}
$$

\sectionbreak

# Constraints

## Regression Constraints

In this paper we use the constraint
\begin{equation}
\sum_{\ell=1}^{k_j}(\delta_{i\ell}^j-\overline{\delta}_{ij})^2=c_j,\label{eq-norm}
\end{equation}
with $\overline{\delta}_{ij}$ the mean of the $\delta_{i\ell}^j$
and with $c_j$ constants. For instance we could choose
$c_j=1$ or $c_j=k_j$ for all $j$. In addition, of course, the 
$\delta_{i\ell}^j$ must satisfy the ordinal constraints. Constraint
\eqref{eq-norm} explicitly excludes the possibility that the
$\delta_{i\ell}^j$ for given $i$ and $j$ are equal for all $\ell$.

Consider the cone of all vectors $\delta$ in $\mathbb{R}^n$ for which
$\delta_i\leq\delta_\ell$ for all $i\neq \ell$ and $e'\delta=0$. The extreme rays of the cone will be the $n-1$ vectors of the form $\alpha e_i+\beta(e-e_i)$ with
$i\neq \ell$. Thus element $i$ is equal to $\alpha$ and all other elements
(including $\delta_\ell$) are equal to $\beta$. Thus we must have $\alpha>\beta$.
The sum of the elements is $\alpha+(n-1)\beta$, which must be zero. 
Thus $\alpha=-(n-1)\beta$, which implies $\beta<0$ and $\alpha>0$.
The extreme rays are positive multiples of the vectors $e_i-n^{-1}e$ with $i\neq\ell$ and the cone is the set of all weighted sums with non-negative weights of these $n-1$ vectors. Here is a small example.
```{r}
a <- -cbind(1, -diag(4))
b <- (diag(5)-(1 / 5))[, -1]
print(a)
print(b)
print(a %*% b)
```
For later reference the sum of squares of the vectors $e_i-n^{-1}e$ is
$(n-1)/n$, so the vectors $\sqrt{n/(n-1)}(e_i-n^{-1}e)$ are the
vectors of length 1 on the extreme rays. 


Suppose $y$ is a vector with $n$ elements. The problem we study is 
to minimixe the sum of squares $\sigma(x):=\|x-y\|^2$ over
$x\in K$ and $x'Jx=1$. Here 

* $K$ is the cone of vectors with $\smash{x_1\leq x_2\leq\cdots\leq x_n}$,
* $J$ is the centering matrix $\smash{J:=I-n^{-1}ee'}$, where
* $e$ is the vector with all $n$ elements equal to one.

Change variables to $\tilde x=Jx$. Then $x'Jx=\tilde x'\tilde x$ and $x\in K$ if and only if $\tilde x\in K$. Also $\tilde x = Jx$
if and only if $x=\tilde x+\beta e$. So the transformed problem is
to minimize
$\|\tilde x+\beta e-y\|^2$ over $\beta$ and $\tilde x\in K$, with
$e'\tilde x=0$ and $\tilde x'\tilde x=1$. The minimizer for $\beta$ is $\overline{y}:=n^{-1}e'y$,
and thus the problem becomes minimizing
$\|\tilde x- Jy\|^2$ over $\tilde x\in K$, $\tilde x'\tilde x=1$, and $e'\tilde x=0$. 

Forget about the constraint $e'\tilde x=0$. If $M$ is the monotone regression operator 
it follows
that the solution for $\tilde x$ is $M(Jy)/\|M(Jy)\|$, which automatically
satisfies $e'\tilde x=0$. Thus the solution of our original problem is 
$$
\hat x=\frac{M(y-\overline{y})}{\|M(y-\overline{y})\|}+\overline{y}.
$$
$M(y-\overline{y})=M(y)-\overline{y}$
and thus $\|M(y-\overline y)\|^2=\|M(y)\|^2+n\overline{y}^2-2\overline{y}e'M(y)$
nor $e'M(y)=e'y=n\overline{y}$.
Thus $\|M(y-\overline y)\|^2=\|M(y)\|^2-n\overline{M(y)}^2$
$$
\hat x=\frac{M(y)-\overline{y}e}{\|M(y-\overline{y})\|}+\overline{y}e=
\frac{1}{\|M(y-\overline{y})\|}M(y)+\left\{1-\frac{1}{\|M(y-\overline{y})\|}\right\}\overline{y}e
$$

## Configuration Constraints

If there are constraints on $X$ and/or $Y$ we use majorization 
theory for restricted MDS( from (@deleeuw_heiser_C_80). Suppose 
$$
\tilde Z:=\begin{bmatrix}\tilde X\\\tilde Y\end{bmatrix}
$$
is the unrestricted Guttman transform. In the majorization step we minimize (or decrease)
\begin{equation}
\text{tr}\ (Z-\tilde Z)'V(Z-\tilde Z)=m\|X-\tilde X\|^2+n\|Y-\tilde Y\|^2-2\ \text{tr}\ (X-\tilde X)'ee'(Y-\tilde Y)\label{eq-partition}
\end{equation}
over $X$ and $Y$ satisfying the constraints.

### Column-Centering

If we require $X'e=0$ then the update becomes $J_n\tilde X$, with
$J_n=I-n^{-1}ee'$ the centering matrix.

This is different from updating $X$ as $J_n\tilde X$ and $Y$
as $\tilde Y-n^{-1}ee'\tilde X$. With this alternative update, in which we
subtract the column-means of $\tilde X$ from both $\tilde X$
and $\tilde Y$, the distances are the same before and after
centering. Thus the distances in all iterations, and consequently the
loss function values, are the same with or without this 
type of centering.

In general it is advantagous to require $e'X=0$ because the third
term in equation \eqref{eq-partition}
becomes independent of $X$, and thus the $(X,Y)$ minimization step
is separable (can be solved for $X$ and $Y$ separately).

### Orthogonality

If we require $X'X=I$, together with $e'X=0$, as in Homogeneity Analysis, 
the solution for $X$ is $KL'$, where $J_n\tilde X=K\Lambda L'$ is a
singular value decomposition. 

### Rank Constraints

One of the main themes in @gifi_B_90 is that the $k_j\times p$ matrices of category quantifications can be constrained to be of rank $r<p$. We will
restrict ourselves in this paper to $r=1$, which defines what Gifi calls
*single quantifications*. If all variables are restricted to have
single quantifications then a Homogeneity Analysis becomes equivalent
to a Principal Component Analysis of the transformed variables. 

In Voronoi Homogeneity Analysis the Voronoi regions of single variables
become parallel strips. The category-points of a variable are on a line through the origin, and the Voronoi lines or planes are parallel and perpendicular to the line with the category-points.

$$
-2\text{tr}\ (X-\tilde X)'ee'(Y-\tilde Y)+n\|Y-\tilde Y\|^2=\text{tr}\ \|(Y-\tilde Y)-n^{-1}ee'(X-\tilde X)\|^2+C
$$
So rank one approximation of $\tilde Y+n^{-1}ee'(X-\tilde X)$. If $X$ is 
column-centered then of course $\tilde Y-ee'\tilde X)$.

### Centroid 

A more complicated restriction is $Y=D^{-1}G'X$, together with $X'X=I$ and $e'X=0$.

This eliminates
$Y$ from the problem and only requires minimization of 
$$
m\|X-\tilde X\|^2+n\|D^{-1}G'X-\tilde Y\|^2-2\text{tr}\ (X-\tilde X)'ee'(D^{-1}G'X-\tilde Y)
$$
$$
-2m\text{tr}\ X'\tilde X+n\text{tr}\ X'GD^{-2}G'X-2n\text{tr}\ X'GD^{-1}\tilde Y
+2\text{tr}\ X'GD^{-1}e_me_n'\tilde X+C
$$
$$
nGD^{-2}G'X-m\tilde X-nGD^{-1}\tilde Y+GD^{-1}e_me_n'\tilde X+e_n\lambda'+XM
$$
Now $e_n'GD^{-1}=e_m'$ and thus
$$
ne_m'D^{-1}G'X-me_n'\tilde X-ne_m'\tilde Y+me_n'\tilde X+n\lambda'=0
$$
$$
e_n\lambda'=-(e_ne_m'D^{-1}G'X-e_ne_m'\tilde Y)
$$
$$
\{nGD^{-1}-e_ne_m'\}\{D^{-1}G'X-\tilde Y\}-(mI_n-GD^{-1}e_me_n')\tilde X=XM
$$

\sectionbreak

# Software

\sectionbreak


# Examples

## Small Example

We start with the small example we used before to illustrate 
HGA.

```{r smallvha, echo = FALSE}
h_small<-vha(small, itmax = 1000, eps = 1e-8, verbose = FALSE)
par(mfrow=c(1, 3))
plotVHA(h_small$xmat, h_small$ymat, ncat = ncat_small, labs = labs_small, pdf = FALSE)
```
Loss is `r formatC(h_small$loss, digits = 10, format = "f")` after `r h_small$itel` iterations. If we compare the plots from the Voronoi
analysis with the star plots from the initial Homogeneity Analysis for this example we see that the algorithm did not have to change much to obtain perfect fit. One thing that does change is that the iterations moves
the category-points "p" and "r" for variable 2 very close to each other.
At the point where we stop the iterations they are still distinguishable,
and thus their perpendicular bisector is still a line in the Voronoi
plot. If we would continue iterating "p" and "r" would coincide
and there would only be two Voronoi regions: one containing category-point
"q" and object-point 2, and one containing all other points.
one containing all other 

## GALO

```{r galovha, echo = FALSE, verbose = FALSE, cache = TRUE}
h_galo <- vha(galo[, 1:4], itmax = 1000, verbose = FALSE)
```
```{r galo_bnds, echo = FALSE}
gmax <- max(rbind(h_galo$xmat, h_galo$ymat))
gmin <- min(rbind(h_galo$xmat, h_galo$ymat))
```
Loss is `r formatC(h_galo$loss, digits = 10, format = "f")` after `r h_galo$itel` iterations.

```{r galo_d_dhat, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Galo D-Dhat Plot",  main =""}
par(pty="s")
plot(h_galo$dmat, h_galo$dhat, col = "RED", pch = 16, xlab = "Distances", ylab = "Disparities")
```

```{r galo_IQ, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Galo IQ",  main =""}
par(pty="s")
plot(voronoi(h_galo$ymat[3:11,], ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(h_galo$ymat[3:11,], as.character(1:9), col = "RED", cex = 1.5)
text(h_galo$xmat, as.character(galo[, 2]), col = "BLUE")
```

```{r galo_ADV, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Galo ADV",  main =""}
par(pty="s")
plot(voronoi(h_galo$ymat[12:18,], ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(h_galo$ymat[12:18,], sort(lab3), col = "RED", cex = 1.5)
text(h_galo$xmat, as.character(galo[, 3]), col = "BLUE", cex = .5)
```
\sectionbreak

# References