---
title: Voronoi Homogeneity Analysis of Categorical Data
author: 
    - name: Jan de Leeuw
      orcid: 0000-0003-1420-1797
      email: jan@deleeuwpdx.net
      affiliation: 
        - name: University of California Los Angeles
          city: Los Angeles
          state: CA
          url: www.ucla.edu
      license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
   pdf:
    fontsize: 12pt
    include-in-header: 
     - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
   html:
    fontsize: 12pt
    include-in-header: 
     - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: TBD
---

```{r loadpackages, echo = FALSE}
suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
suppressPackageStartupMessages(library(MASS, quietly = TRUE))
suppressPackageStartupMessages(library(RSpectra, quietly = TRUE))
suppressPackageStartupMessages(library(deldir, quietly = TRUE))
suppressPackageStartupMessages(library(dismo, quietly = TRUE))
```

```{r sourcecode, echo = FALSE}
source("data/small.R")
source("data/galo.R")
source("data/cetacea.R")
source("previous/mca.R")
source("previous/majorization.R")
source("previous/monotone.R")
source("previous/auxiliaries.R")
source("previous/voronoi.R")
source("previous/plotVHA.R")
```

\sectionbreak 

**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw/voronoi> 

\sectionbreak

# Introduction

## Homogeneity Analysis

*Homogeneity Analysis (HGA)*, a.k.a. *Multiple Correspondence Analysis (MCA)*, can be introduced and implemented in many different ways\footnote{Since PCA and MCA have three-letter
acronyms I added the G of Gifi to the HA of Homogeneity Analysis.}. We refer to
@guttman_41, @tenenhaus_young_85, @bekker_deleeuw_C_88, @gifi_B_90, @greenacre_blasius_06, @leroux_rouanet_10, @husson_josse_saporta_16, and @deleeuw_C_23 for a bouquet of
discussions and interpretations of the equations and algorithms
defining the HGA technique.

A *variable* is a function that maps on a (finite or infinite) domain of *objects*
into a (finite or infinite) codomain of *values*.
In HGA the data are the values of $m$ variables on the same *carrier*, a 
finite subset the domain consisting of $n$ objects. 
Each variable $j$ has a finite number $k_j$ of *categories*.  The categories
constitute the *range* of the variable for the $n$ objects in the carrier. Note that all
variables have the same carrier. Each object maps into a unique category, and each category 
contains a subset of objects from the carrier. We also say that a category has a number of objects that "belong to" the category.

Here is a small example with $n=10$ and $m=3$, taken from @gifi_B_90, page 65.
```{r small, echo = FALSE}
print(small)
```

The key mathematical object in HGA is the
*indicator matrix*, a classical way to code categorical data\footnote{Since the indicator matrix is the core of the Gifi 
system we avoid the irreverent alternative "dummy coding".}.
For a variable $j$ with $k_j$ categories it is an $n\times k_j$ binary
matrix $G_j$ in which each row $i$ contains a single element equal to one,
indicating into which category the object $i$ maps for variable $j$.
The cross product $D_j:=G_j'G_j$ is diagonal, with the *marginal
frequencies* of the categories on the diagonal. For two different variables the crossproduct $C_{j\ell}=G_j'G_\ell$ is the $k_j\times k_\ell$ *cross table* or *contingency table* of the two variables. Here are the three 
indicator matrices of our small example, next to each other.

```{r indi, echo = FALSE}
g1 <- makeIndicator(small[, 1])
g2 <- makeIndicator(small[, 2])
g3 <- makeIndicator(small[, 3])
d1 <- diag(1 / colSums(g1))
d2 <- diag(1 / colSums(g2))
d3 <- diag(1 / colSums(g3))
gg <- noquote(cbind("||", g1, "||", g2, "||", g3, "||"))
colnames(gg) <- c("", "a", "b", "c", "", "p", "q", "r", "", "u", "v", "")
print(gg)
```

We now deviate from the usual Gifi and French approaches (@husson_josse_saporta_16) and define HGA as a form of non-metric multidimensional scaling, using the *stress* loss function (@kruskal_64a, @kruskal_64b). Define\footnote{The symbol $:=$ is used for definitions}
\begin{equation}
\sigma(X,Y_1,\cdots,Y_m,\Delta_1,\cdots,\Delta_m) ):=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}(\delta_{i\ell}^j-d(x_i,y_\ell^j))^2,\label{eq-loss1}
\end{equation}
with $d(x_i,y_\ell^j)$ the Euclidean distance between row $i$ of $X$ and row $\ell$ of $Y_j$.

HGA minimizes \eqref{eq-loss1} over all $n\times p$ configurations $X$ satisfying $X'X=I$, over all $Y_j$, and over all $\Delta_j$ that satisfy $\smash{\delta_{i\ell}^j=0}$ whenever $\smash{g_{i\ell}^j=1}$. All other elements of the $\Delta_j$ are only constrained to be non-negative. 

geometry, disparities

If
\begin{equation}
\sigma_\star(X,Y_1,\cdots,Y_m):=\min_{\Delta_1,\cdots,\Delta_m}\sigma(X,Y_1,\cdots,Y_m,\Delta_1,\cdots,\Delta_m),\label{eq-loss2}
\end{equation}
then
\begin{equation}
\sigma_\star(X,Y_1,\cdots,Y_m)=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}g_{i\ell}^jd^2(x_i,y_\ell^j).\label{eq-loss3}
\end{equation}
Next define 
\begin{equation}
\sigma_{\star\star}(X):=\min_{Y_1,\cdots,Y_m}\sigma_\star(X;Y_1,\cdots,Y_m).\label{eq-loss4}
\end{equation}
The minimum of \eqref{eq-loss3} over the $Y_j$ is attained at the centroids $\hat Y_j:=D_j^{-1}G_j'X$, and if we substitute these optimal $\smash{\hat Y_j}$
in \eqref{eq-loss3} we see that $\sigma_{\star\star}(X)$ is indeed equal to the sum of squares of all lines defining the stars. Minimizing $\sigma_{\star\star}(X)$ over $X'X=I$, and thus minimizing \eqref{eq-loss1} over $(X, Y_j,\Delta_j)$, defines HGA. 
Actually computing the solution turns out to be a straightforward eigenvalue-eigenvector problem. The details are in the publications we referenced at the beginning of this section.

A HGA in two dimensions produces of our small example the following three star plots. Objects
are numbered and in BLUE, categories have bigger labels and are in RED. 
Note that there are ten objects, but the sets of objects $(1,4,7)$, $(8,10)$, and $(6,9)$
have the same *profile* and consequently get mapped into the same object-point.
Thus there are at most six different object-points. The analysis minimizes the sum of squares of the length of all black lines in the three plots, under the condition that the matrix of object scores is column-centered and orthonormal. We can think of the results as a form
of *graph drawing*, where the graph is the union of the $m$ bipartite graphs of all $m$ "belonging to" relations (@deleeuw_michailidis_C_00, @michailidis_deleeuw_A_01).

```{r mca, echo = FALSE, fig.align = "center"}
haux <- mca(cbind(g1, g2, g3), 2)
y1 <- haux$ymat[1:3, ]
y2 <- haux$ymat[4:6, ]
y3 <- haux$ymat[7:8, ]
x <- haux$xmat
par(pty = "s")
par(mfrow = c(1, 3))
plot(rbind(x, y1), type = "n", xlab = "", ylab = "")
text(x, as.character(1:10), col = "BLUE", cex = .8)
text(y1, c("a", "b", "c"), col = "RED", cex = 1.5)
for (i in 1:10) {
  if (small[i, 1] == "a") lines(rbind(x[i,], y1[1, ]))
  if (small[i, 1] == "b") lines(rbind(x[i,], y1[2, ]))
  if (small[i, 1] == "c") lines(rbind(x[i,], y1[3, ]))
}
plot(rbind(x, y2), type = "n", xlab = "", ylab = "")
text(x, as.character(1:10), col = "BLUE", cex = .8)
text(y2, c("p", "q", "r"), col = "RED", cex = 1.5)
for (i in 1:10) {
  if (small[i, 2] == "p") lines(rbind(x[i,], y2[1, ]))
  if (small[i, 2] == "q") lines(rbind(x[i,], y2[2, ]))
  if (small[i, 2] == "r") lines(rbind(x[i,], y2[3, ]))
}
plot(rbind(x, y3), type = "n", xlab = "", ylab = "")
text(x, as.character(1:10), col = "BLUE", cex = .8)
text(y3, c("u", "v"), col = "RED", cex = 1.5)
for (i in 1:10) {
  if (small[i, 3] == "u") lines(rbind(x[i,], y3[1, ]))
  if (small[i, 3] == "v") lines(rbind(x[i,], y3[2, ]))
}
```


## Voronoi Analysis

Although using loss function \eqref{eq-loss1} makes HGA a form of non-metric multidimensional scaling there are some important differences with the more familiar MDS versions minimizing stress. There is no local minima problem in HGA, because it turns out to be a symmetric eigenvalue problem. In HGA, unlike in MDS, we impose the constraint $X'X=I$. Also, and perhaps most importantly, in HGA we do not really expect the minimum loss function value to be zero, or even close
to zero. The loss function measures deviations from a very unrealistic
"model". Zero loss would imply all object-points coincide with 
all category-points they belong to. This
can only be realized, except in trivial cases, if all object and category-points map into a single point, which is prohibited by the orthonormality constraint on the object scores. Thus we get a perfectly determinate and stable solution, at low computational cost, but by imposing strong constraints on the disparities and a rather arbitrary normalization. We can get around this 
seemingly unfortunate situation by admitting we are not fitting or
testing a "model", but we are creating "best" drawings of the bipartite graphs 
that constitute the data.

The lack of a realistic model makes it interesting to continue use the more general loss function \begin{equation}
\sigma(X,Y_1,\cdots,Y_m,\Delta_1,\cdots,\Delta_m) ):=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}^j(\delta_{i\ell}^j-d(x_i,y_\ell^j))^2,\label{eq-loss2}
\end{equation}
with the constraints that
if $g_{i\ell}^j=1$ then merely $\delta_{i\ell}^j\leq\delta_{i\nu}^j$ for all
$\nu=1,\cdots,k_j$. This is significantly weaker than the HGA constraints, which require the $mn$ disparities $\smash{\delta_{i\ell}^j}$ to be zero.
If object $i$ belongs to category $\ell$ of 
variable $j$ we do no longer require the object-point to coincide
with the category-point, but we merely require that the object-point
is at least as close to category-point $\ell$ as to any of the
other $k_j-1$ category-points of that variable.

*Voronoi Homogeneity Analysis* (VHA) of categorical data minimizes
loss function \eqref{eq-loss1} over $X$, over the $Y_j$ and
over the $\Delta_j$ satisfying the relaxed ordinal constraints. It is consequently
another form of non-metric (ordinal) multidimensional scaling. To be 
more precise, we have $m$ linked non-metric multidimensional unfolding
problems for the binary matrices $G_j$, which are "linked" because they share 
the same $X$. 

Because of the different ordinal constraints on the disparities, the plots corresponding with this analysis are different. As in HGA there is a plot for each variable $j$ with both $n$ object-points and $k_j$ category-points. The $k_j$
category-points are the *generators* of $k_j$ *Voronoi regions* that partition the
space. Voronoi region $\smash{\mathcal{V}^j_\ell}$ is the region of space
where the points are at least as close to generator $\smash{y_\ell^j}$ as to any of the other generators of variable $j$. And loss \eqref{eq-loss4} is zero (for a variable) if all object-points are in the "correct"
Voronoi region, i.e. in the Voronoi region of the generator corresponding with the category they belong to. 

One (inefficient, but conceptually easy) way of constructing
Voronoi regions in the plane is to draw the perpendicular bisectors of 
the (hypothetical) lines connecting all pairs of category-points of a variable. In higher dimensions $p>2$ the bisectors are planes of dimension $p-1$. The convex polyhedral regions thus formed, some bounded and some 
unbounded, define the Voronoi regions\footnote{In the plane a Voronoi region is unbounded if and only if its generator is on the boundary of the convex hull of all generators.}.
The literature on properties and computation of Voronoi regions 
is large, and we refer to the textbooks of @okabe_boots_sugihara_chiu_00
and @aurenhammer_klein_lee_13.

Since the category-point itself is obviously in its own Voronoi
region and since Voronoi regions are convex it follows that
if we make stars by connecting category-points with object
points as in HGA then we actually require
each star to be in its Voronoi region. It is no longer 
true that a category-point is the centroid of points of 
the objects belonging to the category. The category-point can even be outside the 
convex hull of the object-points belonging to the category.

The constraint $X'X=I$ from HGA is no longer
imposed directly, but we somehow have to deal with
the fact that in VHA the minimum of loss function
\eqref{eq-loss1} is always zero.
The trivial solution, in which all
points and regions collapse into a single point and all
$\Delta_j$ are zero, is always available and not excluded by the constraints.
But, perhaps more urgently, we have to deal
with the types of trivial solutions familiar from multidimensional
unfolding (@heiser_81, @deleeuw_R_83, @busing_10). For example,
we can collapse all category-points of a variable into a single point and 
put all object-points on a sphere with the category-point
at its center. Then all distances between category and
object-points are the same and we can choose all disparities
equal to that distance value and obtain loss zero. 

\sectionbreak

# Algorithm

We use an *Alternating Least Squares* or *ALS* algorithm, similar to the approach used in many
other non-metric scaling methods, to minimize
loss \eqref{eq-loss5}. For context, see @borg_groenen_05.
We alternate minimization over $\Delta_1,\cdots,\Delta_m$
for given $X$ and $Y_1\cdots Y_m$ with minimization over $X$ and $Y_1\cdots Y_m$ for
given $\Delta_1,\cdots,\Delta_m$. Using superscript $(\mu)$ for the iteration index we can write for an ALS iteration
\begin{subequations}
\begin{align}
(\Delta_1^{(\mu)},\cdots,\Delta_m^{(\mu)})&=\mathop{\text{argmin}}_{\Delta_1,\cdots,\Delta_m}\sigma(X^{(\mu)},Y_1^{(\mu)},\cdots,Y_m^{(\mu)},\Delta_1,\cdots,\Delta_m),\label{eq-als1}\\
(X^{(\mu+1)},Y_1^{(\mu+1)},\cdots,Y_m^{(\mu+1)})&=\mathop{\text{argmin}}_{X,Y_1,\cdots,Y_m}\sigma(X,Y_1,\cdots,Y_m,\Delta_1^{(\mu)},\cdots,\Delta_m^{(\mu)}).\label{eq-als2}
\end{align}
\end{subequations}
The first substep \eqref{eq-als1} of iteration $(\mu)$ solves
a monotone regression problem for a simple partial order, in which
one of $k_j$ elements must be smaller than the other $k_j-1$ elements.
The second substep \eqref{eq-als2} solves a metric multidimensional scaling problem.
This second subproblem cannot be solved completely, because it would need an 
infinite iterative process by itself. Thus the second substep \eqref{eq-als2}
consists of a finite number of *inner iterations* within the
second substep of a single ALS *outer iteration*. 

## Monotone Regression

The $nm$ monotone regressions are applied for each object $i$ and
each variable $j$. Forgetting about all these indices for a 
moment we need to minimize the sum of squares
$\|\delta-d\|^2$ over $\delta$, where one of the elements, 
say $\delta_\nu$, must be smaller than or equal to all others.

This can be solved by using monotone regression with the 
primary approach to ties
(@kruskal_64b, @deleeuw_A_77), treating all $\delta_\ell$
with $\ell\neq\nu$ as ties.

## Majorization

For the inner iterations we use standard *smacof* majorization steps
(@deleeuw_C_77). Because of the structure of the problem it is
advantageous to decompose a majorization step into two substeps:
updating $X$ for $Y_j$ fixed at its current values and updating
the $Y_j$ for $X$ fixed at its current value. This could be
called *Alternating Majorization*.

Let's first update $X$. The majorization equality we use, for current fixed $Y_j$
and a current value $\tilde X$ for $X$, is a version of Cauchy-Schwartz.
\begin{equation}
d_{i\ell}^j(X,Y_j)\geq\frac{1}{d_{i\ell}^j(\tilde X,Y_j)}(x_i-y_\ell^j)'(\tilde x_i- y_\ell^j).
\end{equation}
If we define 
\begin{equation}
b_{i\ell}^j(\tilde X,Y_j):=w_{i\ell}^j\frac{\delta_{i\ell}^j}{d_{i\ell}^j(\tilde X,Y_j)}
\end{equation}
then
\begin{align}
\sigma_j(X,Y_j)\leq C_j&-2\sum_{i=1}^n\sum_{\ell=1}^{k_j}b_{i\ell}^j(\tilde X,Y_j)(x_i-y_\ell^j)'(\tilde x_i- y_\ell^j)\notag\\&+\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}^j(x_i-y_\ell^j)'(x_i-y_\ell^j),\label{eq-majx1}
\end{align}
where $C_j$ does not depend of $X$. The right-hand-side of \eqref{eq-majx1} must be 
minimized over $X$. Use the symbol $\star$ to replace a subscript or 
superscript over which we have summed. The stationary equations are\
\begin{subequations}
\begin{equation}
w_{i\star}^\star x_i=b_{i\star}^\star(\tilde X,Y_j)\tilde x_i-\sum_{j=1}^m\sum_{\ell=1}^{k_j}(b_{i\ell}^j(\tilde X,Y_j)-w_{i\ell}^j)y_\ell^j.
\end{equation}
In the same way we can update $\tilde Y_j$. By symmetry
\begin{equation}
w_{\star\ell}^jy_\ell^j=b_{\star\ell}^j(X,\tilde Y_j)\tilde y_\ell^j-\sum_{i=1}^n(b_{i\ell}^j(X,\tilde Y_j)-w_{i\ell}^j)x_i
\end{equation}
\end{subequations}

 
\sectionbreak

# Constraints

## Regression Constraints

In this paper we use the constraint
\begin{equation}
\sum_{\ell=1}^{k_j}(\delta_{i\ell}^j-\overline{\delta}_{ij})^2=c_j,\label{eq-norm}
\end{equation}
with $\overline{\delta}_{ij}$ the mean of the $\delta_{i\ell}^j$
and with $c_j$ constants. For instance we could choose
$c_j=1$ or $c_j=k_j$ for all $j$. In addition, of course, the 
$\delta_{i\ell}^j$ must satisfy the ordinal constraints. Constraint
\eqref{eq-norm} explicitly excludes the possibility that the
$\delta_{i\ell}^j$ for given $i$ and $j$ are equal for all $\ell$.

Consider the cone of all vectors $\delta$ in $\mathbb{R}^n$ for which
$\delta_i\leq\delta_\ell$ for all $i\neq \ell$ and $e'\delta=0$. The extreme rays of the cone will be the $n-1$ vectors of the form $\alpha e_i+\beta(e-e_i)$ with
$i\neq \ell$. Thus element $i$ is equal to $\alpha$ and all other elements
(including $\delta_\ell$) are equal to $\beta$. Thus we must have $\alpha>\beta$.
The sum of the elements is $\alpha+(n-1)\beta$, which must be zero. 
Thus $\alpha=-(n-1)\beta$, which implies $\beta<0$ and $\alpha>0$.
The extreme rays are positive multiples of the vectors $e_i-n^{-1}e$ with $i\neq\ell$ and the cone is the set of all weighted sums with non-negative weights of these $n-1$ vectors. Here is a small example.
```{r}
a <- -cbind(1, -diag(4))
b <- (diag(5)-(1 / 5))[, -1]
print(a)
print(b)
print(a %*% b)
```
For later reference the sum of squares of the vectors $e_i-n^{-1}e$ is
$(n-1)/n$, so the vectors $\sqrt{n/(n-1)}(e_i-n^{-1}e)$ are the
vectors of length 1 on the extreme rays. 


Suppose $y$ is a vector with $n$ elements. The problem we study is 
to minimixe the sum of squares $\sigma(x):=\|x-y\|^2$ over
$x\in K$ and $x'Jx=1$. Here 

* $K$ is the cone of vectors with $\smash{x_1\leq x_2\leq\cdots\leq x_n}$,
* $J$ is the centering matrix $\smash{J:=I-n^{-1}ee'}$, where
* $e$ is the vector with all $n$ elements equal to one.

Change variables to $\tilde x=Jx$. Then $x'Jx=\tilde x'\tilde x$ and $x\in K$ if and only if $\tilde x\in K$. Also $\tilde x = Jx$
if and only if $x=\tilde x+\beta e$. So the transformed problem is
to minimize
$\|\tilde x+\beta e-y\|^2$ over $\beta$ and $\tilde x\in K$, with
$e'\tilde x=0$ and $\tilde x'\tilde x=1$. The minimizer for $\beta$ is $\overline{y}:=n^{-1}e'y$,
and thus the problem becomes minimizing
$\|\tilde x- Jy\|^2$ over $\tilde x\in K$, $\tilde x'\tilde x=1$, and $e'\tilde x=0$. 

Forget about the constraint $e'\tilde x=0$. If $M$ is the monotone regression operator 
it follows
that the solution for $\tilde x$ is $M(Jy)/\|M(Jy)\|$, which automatically
satisfies $e'\tilde x=0$. Thus the solution of our original problem is 
$$
\hat x=\frac{M(y-\overline{y})}{\|M(y-\overline{y})\|}+\overline{y}.
$$
$M(y-\overline{y})=M(y)-\overline{y}$
and thus $\|M(y-\overline y)\|^2=\|M(y)\|^2+n\overline{y}^2-2\overline{y}e'M(y)$
nor $e'M(y)=e'y=n\overline{y}$.
Thus $\|M(y-\overline y)\|^2=\|M(y)\|^2-n\overline{M(y)}^2$
$$
\hat x=\frac{M(y)-\overline{y}e}{\|M(y-\overline{y})\|}+\overline{y}e=
\frac{1}{\|M(y-\overline{y})\|}M(y)+\left\{1-\frac{1}{\|M(y-\overline{y})\|}\right\}\overline{y}e
$$

## Configuration Constraints

If there are constraints on $X$ and/or $Y$ we use majorization 
theory for restricted MDS( from (@deleeuw_heiser_C_80). Suppose 
$$
\tilde Z:=\begin{bmatrix}\tilde X\\\tilde Y\end{bmatrix}
$$
is the unrestricted Guttman transform. In the majorization step we minimize (or decrease)
\begin{equation}
\text{tr}\ (Z-\tilde Z)'V(Z-\tilde Z)=m\|X-\tilde X\|^2+n\|Y-\tilde Y\|^2-2\ \text{tr}\ (X-\tilde X)'ee'(Y-\tilde Y)\label{eq-partition}
\end{equation}
over $X$ and $Y$ satisfying the constraints.

### Column-Centering

If we require $X'W_re=0$ then the stationarty equations are
$W_r(X-\overline X)=W_re\lambda'$ and thus
$X=\overline X+e\lambda'$ with $\lambda'=-e'W_r\overline X/e'W_re$
or
$$
X=(I-\frac{ee'W_r}{e'W_re})\overline X
$$

This is different from updating $X$ as $J_n\tilde X$ and $Y$
as $\tilde Y-n^{-1}ee'\tilde X$. With this alternative update, in which we
subtract the column-means of $\tilde X$ from both $\tilde X$
and $\tilde Y$, the distances are the same before and after
centering. Thus the distances in all iterations, and consequently the
loss function values, are the same with or without this 
type of centering.

In general it is advantagous to require $e'X=0$ because the third
term in equation \eqref{eq-partition}
becomes independent of $X$, and thus the $(X,Y)$ minimization step
is separable (can be solved for $X$ and $Y$ separately).

### Orthogonality

If we require $X'W_rX=I$ we have ther staionary equations
$\overline X=XM$, with $M$ a symmetric matrix of Lagrange
multipliers. Thus $$X=\text{procrustus}(\overline X):=\overline X(\overline X'W_r\overline X)^{-\frac12}$$.
If we require both $X'W_rX=I$ and $X'W_re=0$ we have
$$X=\text{procrustus}((I-\frac{ee'W_r}{e'W_re})\overline X)$$

### Rank Constraints

One of the main themes in @gifi_B_90 is that the $k_j\times p$ matrices of category quantifications can be constrained to be of rank $r<p$. We will
restrict ourselves in this paper to $r=1$, which defines what Gifi calls
*single quantifications*. If all variables are restricted to have
single quantifications then a HGA becomes equivalent
to a PCA of the transformed variables. 

In VHA the Voronoi regions of single variables
become parallel strips. The category-points of a variable are on a line through the origin, and the Voronoi lines or planes are parallel and perpendicular to the line with the category-points.

$$\text{tr}\ (Y-za')'W_c(Y-za')=-2z'W_cYa+z'W_cz.a'a$$

So rank one approximation of $\tilde Y+n^{-1}ee'(X-\tilde X)$. If $X$ is 
column-centered then of course $\tilde Y-ee'\tilde X)$.

### Centroid 

A more complicated restriction is $Y=D^{-1}G'X$, possibly together with $X'X=I$ and $e'X=0$.
The *centroid constraints* effectively eliminate the $Y_j$ from the optimization 
problem. We have to adapt the majorization method to take the constraints into account.

Consider a single variable $j$ and the corresponding $n\times k_j$ distance
matrix $D_j(X)$. We have $d_{i\ell}(X)=\|t\|$, where
$t:=X'r$ and $r:=e_i-d_l^{-1}g_l$. Now
\begin{subequations}
\begin{align}
d_{i\ell}&=\sqrt{t't}\geq\frac{1}{\sqrt{\tilde t'\tilde t}}\tilde t't=\frac{1}{\sqrt{\tilde t'\tilde t}}\tilde t'X'r=\text{tr}\ X'\frac{r\tilde t'}{\sqrt{\tilde t'\tilde t}},\\
d_{il}^2&=t't=\text{tr}\ X'rr'X,
\end{align}
\end{subequations}
with $\tilde t:=\tilde X'r$.

The loss component for the variable is
$$
\sigma_j(X)=\sum_{i=1}^n\sum_{\ell=1}^{k_j}w_{i\ell}(\delta_{i\ell}-d_{ij}(X))^2
$$

\sectionbreak

# Software

\sectionbreak


# Examples

## Small Example

We start with the small example we used before to illustrate 
HGA.

```{r smallvha, echo = FALSE}
h_small<-vha(small, itmax = 1000, eps = 1e-8, verbose = FALSE)
par(mfrow=c(1, 3))
plotVHA(h_small$xmat, h_small$ymat, ncat = ncat_small, labs = labs_small, pdf = FALSE)
```
Loss is `r formatC(h_small$loss, digits = 10, format = "f")` after `r h_small$itel` iterations. If we compare the plots from the Voronoi
analysis with the star plots from the initial HGA for this example we see that the algorithm did not have to change much to obtain perfect fit. One thing that does change is that the iterations moves
the category-points "p" and "r" for variable 2 very close to each other.
At the point where we stop the iterations they are still distinguishable,
and thus their perpendicular bisector is still a line in the Voronoi
plot. If we would continue iterating "p" and "r" would coincide
and there would only be two Voronoi regions: one containing category-point
"q" and object-point 2, and one containing all other points.
one containing all other 

## Cetacea

## GALO

```{r galovha, echo = FALSE, verbose = FALSE, cache = TRUE}
h_galo <- vha(galo[, 1:4], itmax = 1000, verbose = FALSE)
```
```{r galo_bnds, echo = FALSE}
gmax <- max(rbind(h_galo$xmat, h_galo$ymat))
gmin <- min(rbind(h_galo$xmat, h_galo$ymat))
```
Loss is `r formatC(h_galo$loss, digits = 10, format = "f")` after `r h_galo$itel` iterations.

```{r galo_d_dhat, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Galo D-Dhat Plot",  main =""}
par(pty="s")
plot(h_galo$dmat, h_galo$dhat, col = "RED", pch = 16, xlab = "Distances", ylab = "Disparities")
```

```{r galo_IQ, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Galo IQ",  main =""}
par(pty="s")
plot(voronoi(h_galo$ymat[3:11,], ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(h_galo$ymat[3:11,], as.character(1:9), col = "RED", cex = 1.5)
text(h_galo$xmat, as.character(galo[, 2]), col = "BLUE")
```

```{r galo_ADV, fig.align = "center", fig.width = 10, fig.height = 10, echo = FALSE, fig.cap = "Galo ADV",  main =""}
par(pty="s")
plot(voronoi(h_galo$ymat[12:18,], ext = 1.1 * c(gmin, gmax, gmin, gmax)))
text(h_galo$ymat[12:18,], sort(lab3), col = "RED", cex = 1.5)
text(h_galo$xmat, as.character(galo[, 3]), col = "BLUE", cex = .5)
```

\sectionbreak

# Appendix: Majorized Alternating Least Squares (MALS)

Consider the following problem: minimize the loss function
$$
\sigma(X,Y):=\sum_{i=1}^n\sum_{j=1}^mw_{ij}(\delta_{ij}-d(x_i,y_j))^2
$$
over $X$ in a subset $\mathcal{X}$ of $\mathbb{R}^{n\times p}$ and $Y$ in a subset
$\mathcal{Y}$ of $\mathbb{R}^{m\times p}$. Here $d(x_i,y_j)$ is the Euclidean distance
between row $i$ of $X$ and row $j$ of $Y$.
The weights $w_{ij}$ and the dissimilarities $\delta_{ij}$ are given non-negative
numbers.

@heiser_deleeuw_A_79

The algorithm we propose in this appendix alternates between decreasing loss by 
modifying $X$ for fixed $Y$ and decreasing loss by modifying $Y$ for fixed $X$.
This is the alternating least squares (ALS) component of our algorithm. Majorization
is applied to both the function of $X$ minimized in the first ALS step and to
the function of $Y$ minimized in the second ALS step.

We derive the update formula for $X$, the one for $Y$ follows by symmetry.
First expand stress
$$
\sigma(X,Y)=C-2\ \sum_{i=1}^n\sum_{j=1}^mw_{ij}\delta_{ij}d(x_i,y_j)+\sum_{i=1}^n\sum_{j=1}^mw_{ij}d_{ij}^2(x_i,y_j)
$$
Here $C$ is used as a generic term that does not depend on $X$. So $C$ can mean
different things in different formulas.

Suppose $\tilde X$ and $\tilde Y$ are our current values of $X$ and $Y$. Then by the CS-inequality
$$
d(x_i,\tilde y_j)\geq\frac{1}{d(\tilde x_i,\tilde y_j)}(x_i-\tilde y_j)(\tilde x_i-\tilde y_j)
$$
Define 
$$
\tilde b_{ij}:=w_{ij}\frac{\delta_{ij}}{d(x_i,y_j)}
$$
Then
$$
\sigma(X,\tilde Y)\leq\eta(X,\tilde X,\tilde Y)
$$
with
$$
\eta(X,\tilde X,\tilde Y):=
C-2\ \sum_{i=1}^n\sum_{j=1}^m\tilde b_{ij}(x_i-\tilde y_j)'(\tilde x_i-\tilde y_j)+\sum_{i=1}^n\sum_{j=1}^mw_{ij}(x_i-\tilde y_j)'(x_i-\tilde y_j)
$$

Now
$$
\sum_{i=1}^nx_i'\sum_{j=1}^m\tilde b_{ij}(\tilde x_i-\tilde y_j)=\text{tr}\ X'(\tilde B_r\tilde X-\tilde B\tilde Y)+C
$$
with $\tilde B_r$ the diagonal matrix of row sums of $\tilde B$. Also

$$
\sum_{i=1}^n\sum_{j=1}^mw_{ij}(x_i-\tilde y_j)'(x_i-\tilde y_j)=\text{tr}\ X'W_r X-2\ \text{tr}\ X'W\tilde Y+C
$$
with $W_r$ the diagonal matrix of row sums of $W$. Gathering terms gives
$$
\eta(X,\tilde X, \tilde Y)=\text{tr}\ (X-\overline X)'W_r(X-\overline X)+C
$$
with 
$$
\overline {X}=W_r^{-1}(\tilde B_r\tilde X-(\tilde B-W)\tilde Y)
$$

By symmetry
$$
\eta(Y,\tilde X, \tilde Y)=\text{tr}\ (Y-\overline Y)'W_c(Y-\overline Y)+C
$$
with
$$
\overline {Y}=W_c^{-1}(\tilde B_c\tilde Y-(\tilde B-W)'\tilde X)
$$
using the diagonal matrices $W_c$ and $B_c$ of column sums.

The algorithm to minimize stress is thus as follows. Our current best values in
iteration $k$ are $\smash{X^{(k)}}$ and $\smash{Y^{(k)}}$. 
$$
X^{(k+1)}=\text{proj}_r(W_r^{-1}(\tilde B_r^{(k,k)}X^{(k)}-(B^{(k,k)}-W)Y^{(k)}))
$$
$$
Y^{(k+1)}=\text{proj}_c(W_c^{-1}(\tilde B_c^{(k+1,k)}Y^{(k)}-(B^{(k+1,k)}-W)'X^{(k+1)}))
$$
\sectionbreak

# References