---
title: Vomals -- Voronoi Analysis of Categorical Data
author: 
    - name: Jan de Leeuw
      orcid: 0000-0003-1420-1797
      email: jan@deleeuwpdx.net
      affiliation: 
        - name: University of California Los Angeles
          city: Los Angeles
          state: CA
          url: www.ucla.edu
      license: "CC0"
date: last-modified
date-format: long
bibliography: [mypubs.bib, total.bib]
number-sections: true
pdf-engine: lualatex
keep-tex: true
format:
   pdf:
    fontsize: 12pt
    include-in-header: 
     - preamble.tex
    keep-tex: true
    link-citations: true
    documentclass: scrartcl
    number-sections: true
   html:
    fontsize: 12pt
    include-in-header: 
     - preamble.css
    keep-md: true
    number-sections: true
toc: true
toc-depth: 3
editor: source
papersize: letter
graphics: true
link-citations: true
mainfont: Times New Roman
abstract: TBD
---

```{r loadpackages, echo = FALSE}
suppressPackageStartupMessages(library(knitr, quietly = TRUE))
suppressPackageStartupMessages(library(tinytex, quietly = TRUE))
```

\sectionbreak 

**Note:** This is a working manuscript which will be expanded/updated
frequently. All suggestions for improvement are welcome. All Rmd, tex,
html, pdf, R, and C files are in the public domain. Attribution will be
appreciated, but is not required. The files can be found at
<https://github.com/deleeuw/voronoi> 

\sectionbreak

# Introduction

# Homogeneity Analysis

*Homogeneity analysis*, a.k.a. *Multiple Correspondence
Analysis*, can be introduced in many different ways. We refer to
@guttman_41, @tenenhaus_young_85, @bekker_deleeuw_C_88, @gifi_B_90, @greenacre_91, @leroux_rouanet_10, @deleeuw_C_23 for a bouquet of
discussions and interpretations of the equations and algorithms
defining the technique.

For our purposes a geometric interpretation using *star plots* seems
the most natural one. We have $n$ objects measured by $m$ 
categorical variables. Variable $j$ has $k_j$ categories, which
are the different values the variable can assumes for the $n$ objects.
Suppose the objects are mapped into $n$ points of Euclidean space
$\mathbb{R}^p$. For each variable we can construct an additional
$k_j$ points in the same space by computing for each category
the centroids of all object-points in that category. Thus each
variable defines its own plot. We then draw lines from the $n$ 
object points to "their" category point and we obtain $k_j$ 
"stars". Homogeneity analysis maps the objects
into space in such a way that the stars are as small as possible,
where the size of the star is measured as the sum of squares
of the distances between each category points and "its" object
points (summed over categories and over variables). In order
to avoid the trivial solution in which all objects are mapped
into the same point homogeneity requires that the *configuration*
of object points (an $n\times p$ matrix) is column-centered and
orthonormal.

There are a huge number of variations possible on the same theme.
Instead of the centroids we could use the Weber points to define
our stars. Instead of Euclidean distance we could use any other
distance on $\mathbb{R}^p$. Instead of the stars we could use the 
convex hulls or the minimum spanning trees, or any other way 
to visualize and quantify the compactness of the point clouds of each category. Any one of these choices defines a technique that could be called homogeneity analysis, but in this paper we mean the classical
choice that uses stars, squared Euclidean distance, and orthonormal
object points.

The key mathematical object in homogeneity analysis is the
*indicator matrix*, which is a classical way to code categorical data.
For a variable $j$ with $k_j$ categories it is an $n\times k_j$ binary
matrix $G_j$ in which each row $i$ contains a single element equal to one,
indicating into which category the object $i$ maps for variable $j$.
The cross product $D_j:=G_j'G_j$ is diagonal, with the *marginal
frequencies* of the categories on the diagonal. For two different variables the crossproduct $C_{j\ell}=G_j'G_\ell$ is the $k_j\times k_\ell$ *cross table* or *contingency table* of the two variables.

The $G_j$ can be used as weights in the traditional "stress" multidimensional scaling loss function, defined as
$$
\sigma(X;Y_1,\cdots,Y_m):=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}g_{i\ell}^j(\delta_{i\ell}^j-d(x_i,y_\ell^j))^2.
$$
Here the dissimilarity $\delta_{i\ell}^j=0$ if $g_{i\ell}^j=1$. If
$g_{i\ell}^j=0$ the dissimilarity is arbitrary, because that
particular triple $(i,j,\ell)$ does not contribute to the loss 
function. Since $\delta_{i\ell}^jg_{i\ell}^j=0$ for all $(i,j,\ell)$ we
have
$$
\sigma(X;Y_1,\cdots,Y_m)=\sum_{j=1}^m\sum_{i=1}^n\sum_{\ell=1}^{k_j}g_{i\ell}^j\{d(x_i,y_\ell^j)\}^2.
$$
Define 
$$
\sigma_\star(X):=\min_{Y_1,\cdots,Y_m}\sigma(X;Y_1,\cdots,Y_m).
$$
The minimum over the $Y_j$ is attained at the centroids $\hat Y_j:=D_j^{-1}G_j'X$, and if we substitute these optimal $\hat Y_j$
in ... we see that $\sigma_\star(X)$ is indeed equal to our measure of the total size of all stars. Minimizing $\sigma_\star(X)$ over $X'X=I$
is homogeneity analysis, and computing the solution turns out to be a straightforward eigenvalue-eigenvector problem.

\sectionbreak

# Voronoi Analysis

$$
\sigma(X,Y,\Delta)=\sum_{i=1}^n\sum_{j=1}^m\|\delta _{ij}-d_{ij}(X,Y)\|^2
$$

If $g_{i\ell}^j=1$ then $\{\delta_{ij}\}_\ell\leq\{\delta_{ij}\}_\nu$
for all $\nu=1,\cdots,k_j$.

$$
\delta_{ij}'J\delta_{ij}=1
$$
# Majorization


$$
A_{ij}:=\begin{bmatrix}e_ie_i'&-e_ie_j'\\-e_je_i'&e_je_j'\end{bmatrix}
$$
$$
Z:=\begin{bmatrix}
X\\Y
\end{bmatrix}
$$

$$
d_{ij}(X,Y)=\text{tr}\ Z'A_{ij}Z
$$
$$
V:=\sum_{i=1}^n\sum_{j=1}^m A_{ij}=\begin{bmatrix}\hfill mI&-ee'\\-ee'&\hfill nI\end{bmatrix}
$$
```{r check1, echo = FALSE, eval = FALSE}
m <- 3
n <- 5
library(MASS)
v <- matrix(0, n + m, n + m)
v[1:n, 1:n] <- m * diag(n)
v[n + (1:m), n + (1:m)] <- n * diag(m)
v[1:n, n + (1:m)] <- -1
v[n + (1:m), 1:n] <- -1
v1 <- matrix(0, n + m, n + m)
v1[1:n, 1:n] <- (m / n)
v1[n + (1:m), n + (1:m)] <- (n / m)
v1[1:n, n + (1:m)] <- -1
v1[n + (1:m), 1:n] <- -1
v1 <- v1 / (n + m)
v2 <- matrix(0, n + m, n + m)
v2[1:n, 1:n] <- diag(n) - (1 / n)
v3 <- matrix(0, n + m, n + m)
v3[n + (1:m), n + (1:m)] <- diag(m) - (1 / m)
vv <- (n + m) * v1 + m * v2 + n * v3
print(max(abs(v - vv)))
vin <- ginv(v)
viv <- v1 / (n + m) + v2 / m + v3 / n
print(max(abs(vin - viv)))
```
If we define
\begin{align}
V_1&:=(n+m)^{-1}\begin{bmatrix}\frac{m}{n}E&-E\\-E&\frac{n}{m}E\end{bmatrix},\\
V_2&:=\begin{bmatrix}J_n&0\\0&0\end{bmatrix},\\
V_3&:=\begin{bmatrix}0&0\\0&J_m\end{bmatrix},
\end{align}
then
$$
V=(n+m)V_1+mV_2+nV_3
$$
Matrices $V_1,V_2,$ and $V_3$ are symmetric projectors on orthogonal
subspaces. It follows that $V^+$, the Moore-Penrose inverse3 of $V$,
is
$$
V^+=\frac{1}{n+m}V_1+\frac{1}{m}V_2+\frac{1}{n}V_3
$$
$$
B(X,Y):=\begin{bmatrix}B_{11}(X,Y)&B_{12}(X,Y)\\B_{21}(X,Y)&B_{22}(X,Y)\end{bmatrix}
$$
Now $V_1B(X,Y)=0$ and thus $$V^+B(X,Y)=\frac{1}{m}V_2B(X,Y)+\frac{1}{n}V_3B(X,Y)$$

# The Cone

Consider the cone of all vectors $x$ in $\mathbb{R}^n$ for which
$x_i\leq x_1$ for all $i\neq 1$ and $e'x=0$. The extreme rays of the
cone will be the $n-1$ vectors of the form $\alpha e_i+\beta(e-e_i)$ with
$i\neq 1$. Thus element $i$ is equal to $\alpha$ and all other elements
(including $x_1$) are equal to beta. Thus we must have $\alpha>\beta$.
The sum of the elements is $\alpha+(n-1)\beta$, which must be zero. 
Thus $\alpha=-(n-1)\beta$, which implies $\beta<0$ and $\alpha>0$.
The extreme rays are positive multiples of the vectors $e_i-n^{-1}e$ with $i\neq 1$ and the cone is the set of all weighted sums with non-negative weights of these $n-1$ vectors. Here is a small example.
```{r}
a <- -cbind(1, -diag(4))
b <- (diag(5)-(1 / 5))[, -1]
print(a)
print(b)
print(a %*% b)
```
For later reference the sum of squares of the vectors $e_i-n^{-1}e$ is
$(n-1)/n$, so the vectors $\sqrt{n/(n-1)}(e_i-n^{-1}e)$ are the
vectors of length 1 on the extreme rays. The average of the
extreme rays is 


# Normalized Monotone Regression

Suppose $y$ is a vector with $n$ elements. The problem we study is 
to minimixe the sum of squares $\sigma(x):=\|x-y\|^2$ over
$x\in K$ and $x'Jx=1$. Here 

* $K$ is the cone of vectors with $\smash{x_1\leq x_2\leq\cdots\leq x_n}$,
* $J$ is the centering matrix $\smash{J:=I-n^{-1}ee'}$, where
* $e$ is the vector with all $n$ elements equal to one.

Change variables to $\tilde x=Jx$. Then $x'Jx=\tilde x'\tilde x$ and $x\in K$ if and only if $\tilde x\in K$. Also $\tilde x = Jx$
if and only if $x=\tilde x+\beta e$. So the transformed problem is
to minimize
$\|\tilde x+\beta e-y\|^2$ over $\beta$ and $\tilde x\in K$, with
$e'\tilde x=0$ and $\tilde x'\tilde x=1$. The minimizer for $\beta$ is $\overline{y}:=n^{-1}e'y$,
and thus the problem becomes minimizing
$\|\tilde x- Jy\|^2$ over $\tilde x\in K$, $\tilde x'\tilde x=1$, and $e'\tilde x=0$. 

Forget about the constraint $e'\tilde x=0$. If $M$ is the monotone regression operator 
it follows
that the solution for $\tilde x$ is $M(Jy)/\|M(Jy)\|$, which automatically
satisfies $e'\tilde x=0$. Thus the solution of our original problem is 
$$
\hat x=\frac{M(y-\overline{y})}{\|M(y-\overline{y})\|}+\overline{y}.
$$
$M(y-\overline{y})=M(y)-\overline{y}$
and thus $\|M(y-\overline y)\|^2=\|M(y)\|^2+n\overline{y}^2-2\overline{y}e'M(y)$
nor $e'M(y)=e'y=n\overline{y}$.
Thus $\|M(y-\overline y)\|^2=\|M(y)\|^2-n\overline{M(y)}^2$
$$
\hat x=\frac{M(y)-\overline{y}e}{\|M(y-\overline{y})\|}+\overline{y}e=
\frac{1}{\|M(y-\overline{y})\|}M(y)+\left\{1-\frac{1}{\|M(y-\overline{y})\|}\right\}\overline{y}e
$$



\sectionbreak

# References